
This coding file conducts imputation on the environmental variables. Note that we are only imputing environmental variables.

```{r}
library(dplyr)
library(tableone)
library(kableExtra)
library(missForest)
library(doParallel)
```

load datasets
```{r}
df <- readRDS()

envraw <- read.csv()
```

restrict to visits with valid lung funciton data (used to check # of missing)
```{r}
df <- df %>% filter(!is.na(v2ht_starttime))
```



#1. testing missing
get number of missing for all variables
```{r}
all_vars <- names(df)

missing_df <- data.frame(
  variable = all_vars,
  n_missing = sapply(df[all_vars], function(x) sum(is.na(x))),
  pct_missing = sapply(df[all_vars], function(x) mean(is.na(x)) * 100)
)
```

get number of missing in env dataset
```{r}
all_vars <- names(envraw)
missing_df_env <- data.frame(
  variable = all_vars,
  n_missing = sapply(envraw[all_vars], function(x) sum(is.na(x))),
  pct_missing = sapply(envraw[all_vars], function(x) mean(is.na(x)) * 100)
)
```

dates where UFP data is missing
```{r}
ufp_vars <- c("PM25_A_LfU_UFP", "PM25_A_LfU_UFP", "BC_880_A_LfU_UFP", "Temp_A_LfU_UFP", "Druck_A_LfU_UFP", "Feuchte_A_LfU_UFP", "NI_A_LfU_UFP")
```

```{r}
mask_missing_any <- rowSums(is.na(envraw[ufp_vars])) > 0
bad_dates_any    <- unique(envraw$date_neu[mask_missing_any])
bad_dates_any    <- bad_dates_any[!is.na(bad_dates_any)]  # drop NA dates if present
parts <- do.call(rbind, strsplit(bad_dates_any, "-", fixed = TRUE))
bad_dates_any <- bad_dates_any[order(match(parts[,2], month.name),
                                     as.integer(parts[,1]))]
print(bad_dates_any)
```




#2. prepare the dataset for imputation

```{r}
X <- envraw

#make sure the date columns are right
X$date_neu <- as.Date(X$date_neu, format = "%d-%B-%Y")
#need the time zone to ensure daylight savings is accounted for
X$datetime <- as.POSIXct(paste(X$date_neu, X$Hour), format = "%Y-%m-%d %H", tz = "UTC")
```

```{r}
X$time_num   <- as.numeric(X$datetime)   # seconds since 1970
X$Hour       <- as.integer(format(X$datetime, "%H"))
X$doy        <- as.integer(format(X$datetime, "%j"))   # 1–365
X$month      <- as.integer(format(X$datetime, "%m"))
X$date_neu <- NULL
X$datetime <- NULL
```

```{r}
all_vars <- names(X)
missing_df_env <- data.frame(
  variable = all_vars,
  n_missing = sapply(X[all_vars], function(x) sum(is.na(x))),
  pct_missing = sapply(X[all_vars], function(x) mean(is.na(x)) * 100)
)
```

check how variables will be imputed
```{r}
#column types and missingness
col_class <- vapply(X, function(z) class(z)[1], character(1))
n_mis     <- vapply(X, function(z) sum(is.na(z)), integer(1))
pct_mis   <- round(100 * n_mis / nrow(X), 2)

# missForest: numeric/integer -> regression; factor/ordered/character -> classification
is_num_like   <- col_class %in% c("numeric","integer")
is_factor_like<- col_class %in% c("factor","ordered","character","logical")

method <- ifelse(is_num_like, "regression",
          ifelse(is_factor_like, "classification", "UNSUPPORTED (check)"))

# Build a data.frame summary
imp_plan <- data.frame(
  variable   = names(X),
  class      = col_class,
  n_missing  = n_mis,
  pct_missing= pct_mis,
  method     = method,
  will_impute= n_mis > 0 & method != "UNSUPPORTED (check)",
  row.names  = NULL
)
```

#3. Missing Forest with 100 trees
runtime: 3 minutes
```{r}
no_cores <- detectCores() - 2     #leave 2 cores left  
cl <- makeCluster(no_cores)  
registerDoParallel(cl)

set.seed(809)

res_100tree <- missForest(X, 
                    verbose = TRUE, 
                    ntree = 100, 
                    parallelize = "forests",
                    variablewise = TRUE) 

imp_100tree <- res_100tree$ximp
```



#report error per variable
```{r}
# Per-variable OOB errors (NRMSE for numeric variables)
err_100tree <- res_100tree$OOBerror
rmse <- sqrt(err_100tree)
sd_vec <- sapply(X, sd, na.rm = TRUE)
nrmseerr_100tree <- rmse / sd_vec

names(nrmseerr_100tree) <- colnames(X)

nrmsetable_100tree <- data.frame(
  variable = names(nrmseerr_100tree),
  NRMSE    = round(nrmseerr_100tree, 3),
  row.names = NULL,
  stringsAsFactors = FALSE
)

print(nrmsetable_100tree)
```


#4. Testing IssForest with 500 trees
runtime: 11 minutes
```{r}
no_cores <- detectCores() - 2     #leave 2 cores left  
cl <- makeCluster(no_cores)  
registerDoParallel(cl)

set.seed(809)

res_500tree <- missForest(X, 
                    verbose = TRUE, 
                    ntree = 500, 
                    parallelize = "forests",
                    variablewise = TRUE) 

imp_500tree <- res_500tree$ximp

stopCluster(cl)
registerDoSEQ()
```

report error per variable
```{r}
# Per-variable OOB errors (NRMSE for numeric variables)
err_500tree <- res_500tree$OOBerror
rmse <- sqrt(err_500tree)
sd_vec <- sapply(X, sd, na.rm = TRUE)
nrmseerr_500tree <- rmse / sd_vec

names(nrmseerr_500tree) <- colnames(X)

nrmsetable_500tree <- data.frame(
  variable = names(nrmseerr_500tree),
  NRMSE    = round(nrmseerr_500tree, 3),
  row.names = NULL,
  stringsAsFactors = FALSE
)

print(nrmsetable_500tree)
```


#5. compare complete case to the random forest imputation
- results are pretty similar, perhaps slighty better for 500 forests
```{r}
#complete case
desc_cc <- CreateTableOne(data = X) 
desc_cc <- print(desc_cc)

#imputed ds
desc_imp_100 <- CreateTableOne(data = imp_100tree)
desc_imp_100 <- print(desc_imp_100)

#imputed ds 2 (more trees)
desc_imp_500 <- CreateTableOne(data = imp_500tree)
desc_imp_500 <- print(desc_imp_500)

#cbind and rename colnames 
desc <- cbind(desc_cc, desc_imp_100, desc_imp_500)
colnames(desc) <- c("Complete case", "100 trees", "500 trees")

kable(desc, format = "markdown", caption = "Table 1: Comparison between complete-case and imputed dataset (parallelize = 'forests'")
```


#6. prepare the imputed dataset
```{r}
imp_env_initial <- imp_500tree
```

correct date information
```{r}
imp_env_initial$datetime <- as.POSIXct(
  imp_env_initial$time_num,
  origin = "1970-01-01",
  tz = "UTC"
)
```

add UFP data variables
- temperature, Pm2.5, Pm10, black carbon, humidity, precipitation
```{r}
# Keep only datetime + selected UFP variables
imp_env <- imp_env_initial %>%
  transmute(
    datetime,
    env_temp_ufp     = Temp_A_LfU_UFP,
    env_PM25_ufp     = PM25_A_LfU_UFP,
    env_PM10_ufp     = PM10_A_LfU_UFP,
    env_humidity_ufp = Feuchte_A_LfU_UFP,
    env_precip_ufp   = NI_A_LfU_UFP,
    env_bc_880_ufp   = BC_880_A_LfU_UFP
  )
```

confirm number of missing is 0
```{r}
all_vars <- names(imp_env)
missing_df_env <- data.frame(
  variable = all_vars,
  n_missing = sapply(imp_env[all_vars], function(x) sum(is.na(x))),
  pct_missing = sapply(imp_env[all_vars], function(x) mean(is.na(x)) * 100)
)
```


#save imputed dataset
save as csv for reference, but also save as rds to converse datetime
```{r}
write.csv(imp_env)
saveRDS(imp_env)
```

