
This coding file updates all the variables based on the data dictionary and puts those most relevant into a dataset called "data"

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(rspiro)
library(slider)
library(zoo)
```

#1 load datasets
```{r}
df_original <- read.csv()
df_householdcomposition <- read.csv()
```

joint datasets
```{r}
df_original <- full_join(df_original, df_householdcomposition, by = "zz_nr")
```

create dataset to input cleaned variables
```{r}
data <- data.frame(ID = as.factor(df_original$zz_id))
```


#2 basic transformations/feature engineering
v2x: visit information
v2i : V2 interview
v2it: transformed variable from the V2 interview
pt: education variables
v2h: lung function protocol
v2ht: transformed variable from the lung function protocol
ATC: medication codes

visit number, date, and time
```{r}
data$v2x_visit <- df_original$v2t_visit

data$v2h_examdate  <- as.POSIXct(df_original$v2cuntdat, format = "%d%b%Y:%H:%M:%S", tz = "UTC")
data$v2ht_starttime <- as.POSIXct(df_original$v2h_23_1, format = "%d%b%Y:%H:%M:%S", tz = "UTC")
data$v2ht_endtime   <- as.POSIXct(df_original$v2h_23_2, format = "%d%b%Y:%H:%M:%S", tz = "UTC")

data$v2ht_starttime <- as.POSIXct(
  paste(as.Date(data$v2h_examdate), format(data$v2ht_starttime, "%H:%M:%S")),
  format = "%Y-%m-%d %H:%M:%S", tz = "UTC"
)

data$v2ht_endtime <- as.POSIXct(
  paste(as.Date(data$v2h_examdate), format(data$v2ht_endtime, "%H:%M:%S")),
  format = "%Y-%m-%d %H:%M:%S", tz = "UTC"
)

# Create month factor
data$month <- as.factor(format(data$v2ht_starttime, "%m"))
```



demographics
```{r}
data$v2it_sex <- as.factor(as.character(df_original$v2csex)) # 1/2 male/female
data$v2it_age <- df_original$v2tb_alteru
data$v2it_borngermany <- factor(ifelse(df_original$v2dt == 1, "1",
                                       ifelse(df_original$v2dt == 2, "0", NA))) # 1/0 yes/no
data$v2i_panel <- as.factor((df_original$v2tb_panel)) #COPD/Hypertonie/T2DM
data$v2i_marital <- as.factor((df_original$v2c011)) #1/2/3/4 married/single/divorced/widowed
data$v2it_livepartner <- factor(ifelse(df_original$v2c012 == 1, "1",
                                       ifelse(df_original$v2c012 == 2, "0", NA))) # 1/0 yes/no

#education
data$pt_eduqual<- as.factor((df_original$ptberufb)) 
#1 = no qualification, 2 = Vocational school / Apprenticeship, 3 = technician school / master school, 4 = Engineering school / Polytechnic, 5 = technical college / university 
data$pt_eduyears <- df_original$ptedyrs #8-17 years
```


employment
```{r}
# 1 = Registered as unemployed, 2 = Housewife/husband, 3 = In training or retraining, 5 = On maternity/parental leave, other leave of absence, 6 = In the non-working phase of partial retirement, 4 = Retired 
data$v2i_unemploymenttype <- replace(df_original$v2c017, df_original$v2c017 < 0 , NA) 

# 1 = Yes, full-time, 2 = Yes, regularly employed part-time, 3 = Yes, employed on a minimal or irregular basis, 4 = No
data$v2i_employmenttype <- replace(df_original$v2c016, df_original$v2c016 < 0 , NA) 

# 1 = full-time, 2 = part-time, 3 = retired, 4 = unemployed (one missing value, where someone reported not working, but did not indicate if they were retired)
data$v2it_employment <- dplyr::case_when(
  # Employed (employmenttype ≠ 4)
  data$v2i_employmenttype == 1 ~ 1,
  data$v2i_employmenttype %in% c(2, 3) ~ 2,

  # Unemployed (employmenttype = 4)
  data$v2i_employmenttype == 4 & data$v2i_unemploymenttype %in% c(4, 6) ~ 3,
  #assume this participant with missing data is retired bc they are well over 65
  data$v2i_employmenttype == 4 & is.na(data$v2i_unemploymenttype) ~ 3, 
  data$v2i_employmenttype == 4 & data$v2i_unemploymenttype %in% c(1, 2, 3, 5) ~ 4,

  TRUE ~ NA_real_
)

data$v2i_employmenttype <- as.factor(as.character(data$v2i_employmenttype))
data$v2i_unemploymenttype <- as.factor(as.character(data$v2i_unemploymenttype))
data$v2it_employment <- as.factor(as.character(data$v2it_employment))
```


calculating equivalized income (from modified OECD scale)
first adult: 1, all others over 14: 0.5, children under 14: 0.3
```{r}
data$v2it_income <- replace(df_original$v2a1_3, df_original$v2a1_3 >= 99 , NA) 
data$v2it_income <- as.factor(as.character(data$v2it_income))
median_values <- c("1" = 250, "2" = 625, "3" = 875, "4" = 1250, "5" = 1750, "6" = 2250,
                   "7" = 2750, "8" = 3250, "9" = 3750, "10" = 4250, "11" = 4750,
                   "12" = 5250, "13" = 5500, "14" = 6500 #can change this value, group is over 6000
)

data$v2it_income <- as.numeric(median_values[data$v2it_income])

data$v2it_equivalizedincome <- ifelse(
  df_original$v2a1_1 == 1,
  data$v2it_income,
  round(
    data$v2it_income /
      (1 + 0.5 * (df_original$v2a1_2_1 + df_original$v2a1_2_4 - 1) +
         0.3 * (df_original$v2a1_2_2 + df_original$v2a1_2_3)), 0)
)

#equivalized income based on categories (50%, 100%, 150%, 200%, and over 200% equivalized income)
data$v2it_equivalizedincomecategories <- cut(
  data$v2it_equivalizedincome,
  breaks = c(-Inf, 625, 1250, 1875, 2500, Inf),
  labels = c(1, 2, 3, 4, 5),
  right = FALSE
)

#poverty indicator, using median equivalized income, or 625
data$v2it_povertyindicator <- factor(
  ifelse(data$v2it_equivalizedincome < 625, 1, 0),
  levels = c(1, 0)
)
```

behaviors
```{r}
#smoking
# 1/0 yes/no
data$v2it_currentsmoking <- factor(ifelse(df_original$v2c068 == 1, "1",
                                       ifelse(df_original$v2c068 == 2, "0", NA)))
# 1/2 regularly/occasionally
data$v2i_smokingfrequency <- replace(df_original$v2c070, df_original$v2c070 < 0 , NA)
data$v2i_smokingfrequency <- as.factor(as.character(data$v2i_smokingfrequency))
data$v2i_numcigarettes <- replace(df_original$v2c071, df_original$v2c071 < 0 , NA)
# 0/1/2  no/yes (regularly)/yes (occasionally)
data$v2it_eversmoked <- factor(ifelse(df_original$v2c073 == 3, "0", df_original$v2c073))
# 1/0 yes/no
data$v2it_othertobacco <- factor(ifelse(df_original$v2c077a == 1, "1",
                                        ifelse(df_original$v2c077a == 2, "0", NA)))
# 1/0 yes/no
data$v2it_secondhandsmoke <- factor(ifelse(df_original$v2_rauchen1 == 1, "1",
                                           ifelse(df_original$v2_rauchen1 == 2, "0", NA)))

# new variable for current smoker / previous smoker / non-smoker (1/2/0)
data$v2it_smoking <- factor(ifelse(data$v2it_currentsmoking == 1, 1,
                                   ifelse(data$v2it_eversmoked %in% c(1, 2), 2,
                                          ifelse(data$v2it_eversmoked == 0, 0, NA)))
)

#physical activity
# 1 = Regularly more than 2 hours per week, 2 = Regularly 1 to 2 hours per week, 3 = Less than 1 hour per week, 4 = No sporting activity in winter
data$v2i_wintersports <- as.factor(as.character(df_original$v2c094))
data$v2i_summersports <- as.factor(as.character(df_original$v2c095))
```

exposures 
the data dictionary says 2 = no, but assuming here that 0 = np
```{r}
#1 = stove/burner/heat source outside the dwelling, 2 = single stove/burner in the dwelling 
data$v2i_homeheating <- as.factor(as.character(df_original$v2umw1a))
# 1/0 yes/no
data$v2it_gascooking <- factor(ifelse(df_original$v2umw2 == 1, "1",
                                           ifelse(df_original$v2umw2 == 2, "0", NA)))
# 1/0 yes/no
data$v2it_coolhome <- factor(ifelse(df_original$v2umw3  == 1, "1",
                                           ifelse(df_original$v2umw3  == 2, "0", NA)))
# 1 = Yes, I use it regularly, 2 = Yes, I use it only occasionally, 3 = Yes, but I don't use it , 4 = No
data$v2it_ac <- factor(ifelse(df_original$v2umw4  %in% c(1,2,3), "1",
                                           ifelse(df_original$v2umw4  == 4, "0", NA)))

# a = gas, b = oil, c = electricity, d = wood, e = coal, coke, briquettes
data$v2it_heatingsource <- as.factor(as.character(
  ifelse(df_original$v2umw1b_a == 1, "a",
  ifelse(df_original$v2umw1b_b == 1, "b",
  ifelse(df_original$v2umw1b_c == 1, "c",
  ifelse(df_original$v2umw1b_d == 1, "d",
  ifelse(df_original$v2umw1b_e == 1, "e", NA)))))
))

# occupational exposures. 1/0 yes/no
data$v2i_sunwork        <- factor(replace(df_original$v2arb10_a, df_original$v2arb10_a < 0, NA))
data$v2i_directcoldwork <- factor(replace(df_original$v2arb10_b, df_original$v2arb10_b < 0, NA))
data$v2i_acwork         <- factor(replace(df_original$v2arb10_c, df_original$v2arb10_c < 0, NA))
data$v2i_heatwork       <- factor(replace(df_original$v2arb10_d, df_original$v2arb10_d < 0, NA))
data$v2i_coldwork       <- factor(replace(df_original$v2arb10_e, df_original$v2arb10_e < 0, NA))
data$v2i_moisturework   <- factor(replace(df_original$v2arb10_f, df_original$v2arb10_f < 0, NA))
data$v2i_uniformwork    <- factor(replace(df_original$v2arb10_g, df_original$v2arb10_g < 0, NA))
```

health status
```{r}
# self-rated health
# 1 = very good , 2 = good, 3 = less good , 4 = bad 
data$v2i_selfhealth <- as.factor(as.character(df_original$v2c033))
# 1 = Better, 2 = Worse, 3 = No change, 4 = I don't know 
data$v2i_selfhealthcompared <- as.factor(as.character(df_original$v2c033a))


#comorbidities (1/0 yes/no)
data$v2it_diabetes <- factor(ifelse(df_original$v2c040 == 1, "1",
                                       ifelse(df_original$v2c040 == 2, "0", NA)))
data$v2it_diabetesphysician <- factor(ifelse(df_original$v2c041 == 1, "1",
                                       ifelse(df_original$v2c041 == 2, "0", NA)))

data$v2it_hypertension <- factor(ifelse(df_original$v2c054 == 1, "1",
                                       ifelse(df_original$v2c054 == 2, "0", NA)))
data$v2it_hypertensionphysician <- factor(ifelse(df_original$v2c056 == 1, "1",
                                       ifelse(df_original$v2c056 == 2, "0", NA)))

#creating an asthma binary category, as there is missingness between cured and not cured
data$v2it_everasthma <- factor(ifelse(df_original$v2da13aa == 1, "1",
                                       ifelse(df_original$v2da13aa == 2, "0", NA)))
data$v2it_asthmacured <- factor(ifelse(df_original$v2da13b == 1, "1",
                                       ifelse(df_original$v2da13b == 2, "0", NA)))
# 0/1/2 never asthma/current asthma/previous asthma, cured
data$v2it_asthma <- with(data, case_when(
  v2it_everasthma == 0 ~ "0",
  v2it_everasthma == 1 & v2it_asthmacured == 1 ~ "2",  # cured
  v2it_everasthma == 1 & v2it_asthmacured == 0 ~ "1",  # not cured
  TRUE ~ NA_character_
))

data$v2it_asthma <- factor(data$v2it_asthma, levels = c("0", "1", "2"))

data$v2it_sleepapnea <- factor(ifelse(df_original$v2c100b3 == 1, "1",
                                       ifelse(df_original$v2c100b3 == 2, "0", NA)))
data$v2it_hayfever <- factor(ifelse(df_original$v2da8aa == 1, "1",
                                       ifelse(df_original$v2da8aa == 2, "0", NA)))
data$v2it_coughsymptoms <- factor(ifelse(df_original$v2ba1 %in% c(1,2), "1",
                                       ifelse(df_original$v2ba1 == 3, "0", NA)))
data$v2it_sputumsymptoms <- factor(ifelse(df_original$v2ba2 %in% c(1,2), "1",
                                       ifelse(df_original$v2ba2 == 3, "0", NA)))
data$v2it_copd <- factor(ifelse(df_original$v2ba4a == 1, "1",
                                       ifelse(df_original$v2ba4a == 2, "0", NA)))
data$v2it_emphysema <- factor(ifelse(df_original$v2ba4b == 1, "1",
                                       ifelse(df_original$v2ba4b == 2, "0", NA)))
```

anthropocentric measurements 
correcting to include decimal places, cutoffs for obesity and whr
```{r}
#adding the before decimal place and after decimal place variables together
data$v2it_height <- df_original$v2b_grosv + (df_original$v2b_grosn / 10)
data$v2it_heightmeters <- data$v2it_height/100
data$v2it_weight <- df_original$v2b_gewiv + (df_original$v2b_gewin / 10)
data$v2i_bmi <- replace(df_original$v2tbmi, df_original$v2tbmi < 0 , NA) 

#Waist and hip circumference is not auto-populated for all of the visits like the other variables. automatically enter it for the other columns
data$v2it_waist <- replace(df_original$v2b_tumfv,df_original$v2b_tumfv < 0, NA) 
data$v2it_waist <- data$v2it_waist + (df_original$v2b_tumfn / 10)
data$v2it_hip <- replace(df_original$v2b_humfv, df_original$v2b_humfv < 0 , NA) 
data$v2it_hip <- data$v2it_hip + (df_original$v2b_humfn / 10)

data <- data %>%
  group_by(ID) %>%
  mutate(
    v2it_waist = first(na.omit(v2it_waist)),
    v2it_hip   = first(na.omit(v2it_hip))
  ) %>%
  ungroup()

#creating new variable for whr -- provided value has only one decimal place
data$v2it_whr <- round(data$v2it_waist / data$v2it_hip, 3)


# 0/1/2/3 underweight/normal/overweight/obese
data$v2it_bmiclass <- cut(
  data$v2i_bmi,
  breaks = c(-Inf, 18.5, 25, 30, Inf),
  labels = c("0", "1", "2", "3"),
  right = FALSE
)
data$v2it_bmiclass <- as.factor(as.character(data$v2it_bmiclass))

# 0/1 low risk/high risk for abdominal obesity
data$v2it_whrclass <- with(
  data,
  ifelse(
    v2it_sex == 1,
    ifelse(v2it_whr < 1.0,  "0", "1"), # men
    ifelse(v2it_whr < 0.85, "0", "1")  # women
  )
)
data$v2it_whrclass <- as.factor(as.character(data$v2it_whrclass))
```


creating new columns based on medication classifications, make them factors 
```{r}
atc_codes <- c("ATC_R03","ATC_R06","ATC_S01G","ATC_R01A",
               "ATC_A10","ATC_C02","ATC_C03","ATC_C07",
               "ATC_C08","ATC_C09")

for(code in atc_codes){
  data[[code]] <- factor(
    as.integer(rowSums(
      df_original[ , startsWith(names(df_original), code)] == 1,
      na.rm = TRUE) > 0),
    levels = c(0,1),
    labels = c("0","1")
  )
}
```



#3. lung function variables

protocol variables
```{r}
#protocol variables
# 1 = Worse than usual , 2 = As usual, 3 = Better than usual 
data$v2h_breathingquality <- as.factor(as.character(df_original$v2h_03))
# 1/0 yes/no
data$v2ht_rhinitissymptoms <- factor(ifelse(df_original$v2h_04 == 1, "1",
                                       ifelse(df_original$v2h_04 == 2, "0", NA)))
# 1/0 yes/no
data$v2ht_respinfection <- factor(ifelse(df_original$v2h_05 == 1, "1",
                                       ifelse(df_original$v2h_05 == 2, "0", NA)))
data$v2ht_respinfectioncured <- replace(df_original$v2h_06, df_original$v2h_06 < 0 , NA) 
# 1 = Excellent, 2 = Very good, 3 = Good, 4 = Less good, 5 = Bad 
data$v2ht_healthtoday <- replace(df_original$v2h_12b, df_original$v2h_12b < 0 , NA) 
data$v2ht_healthtoday <- as.factor(as.character(data$v2ht_healthtoday))

# 1 = normal, 2 = reduced
data$v2ht_cooperation <- replace(df_original$v2h_13, df_original$v2h_13 < 0 , NA) 
data$v2ht_cooperation <- as.factor(as.character(data$v2ht_cooperation))
```

lung function variables
```{r}
data$v2h_FVC <- df_original$v2h_14_1m
data$v2h_FVCpct <- df_original$v2h_14_1p
data$v2h_FEV1 <- df_original$v2h_14_2m
data$v2h_FEV1pct <- df_original$v2h_14_2p
data$v2h_FEV1FVC <- df_original$v2h_14_3m

#transformation from L to mL for easier interpretation
data$v2h_FEV1_mL <- data$v2h_FEV1 * 1000
data$v2h_FVC_mL  <- data$v2h_FVC  * 1000
```

recalculate percent, calculate z-scores
```{r}
#recalculating % predicted for comparison
data[, c("v2ht_FEV1pct", "v2ht_FVCpct")] <- pctpred_GLI(
  age      = data$v2it_age,
  height   = data$v2it_heightmeters,
  gender   = data$v2it_sex,
  ethnicity = 1,          # all Caucasian
  FEV1     = data$v2h_FEV1,
  FVC      = data$v2h_FVC
)


data$v2ht_FEV1pct <- as.numeric(as.character(data$v2ht_FEV1pct))
data$v2ht_FVCpct  <- as.numeric(as.character(data$v2ht_FVCpct))

#calculating z-scores
data[, c("v2ht_FEV1z", "v2ht_FVCz")] <- zscore_GLI(
  age       = data$v2it_age,
  height    = data$v2it_heightmeters,
  gender    = data$v2it_sex,
  ethnicity = 1,          # all Caucasian
  FEV1      = data$v2h_FEV1,
  FVC       = data$v2h_FVC
)

data$v2ht_FEV1z <- as.numeric(as.character(data$v2ht_FEV1z))
data$v2ht_FVCz  <- as.numeric(as.character(data$v2ht_FVCz))
```


convert FVC and FEV1 measures to mL for easier interpretation
```{r}
data <- data %>%
  mutate(
    v2h_FEV1_mL = v2h_FEV1 * 1000,
    v2h_FVC_mL  = v2h_FVC  * 1000
  )
```



#4. exclude participants
remove participants that were excluded
- note that we do not exlcude for visits that are missing spirometry data here, as participant 10 is missing spriometry data for baseline, yet we need the other baseline measures for descriptive calculations
```{r}
data$v2i_ausschluss <- df_original$v2tb_ausschluss

data <- data %>%
  filter(v2i_ausschluss == 0)
```


#5. save dataset
save as csv for reference, but also save as rds to conserve factors
```{r}
write.csv(data)
saveRDS(data)
```





#6. adding in environmental variables
two versions: imputed environmental data and non-imputed environmental data
- from non-imputed data also create complete case datasets

load datasets 
```{r}
df <- readRDS("")

envraw <- read.csv("")
envimp <- readRDS("")
```

restrict to visits with valid lung function data
- not done previously because one participant has invalid lung function data for the baseline visit, but we still need the variables for descriptive analyses
```{r}
df <- df %>% filter(!is.na(v2ht_starttime))
```



#6.1. imputed environmental data

create temporary dataset
- update the date column and datetime col (use UTC without time jumps to match the variables)
```{r}
envimp_data <- data.frame(
  datetime = as.POSIXct(envimp$datetime, format = "%Y-%m-%d %H", tz = "UTC")
)

envimp_data$env_temp <- envimp$env_temp_ufp
envimp_data$env_PM25 <- envimp$env_PM25_ufp
envimp_data$env_PM10 <- envimp$env_PM10_ufp
envimp_data$env_rel_humidity <- envimp$env_humidity_ufp
envimp_data$env_precip <- envimp$env_precip_ufp
envimp_data$env_bc_880 <- envimp$env_bc_880_ufp

#calculate absolute humidity
envimp_data$env_abs_humidity <-
  (6.112 * exp((17.67 * envimp_data$env_temp) /
               (envimp_data$env_temp + 243.5)) *
     (envimp_data$env_rel_humidity / 100) * 
     216.7) / 
  (273.15 + envimp_data$env_temp)

# Ensure data are ordered
envimp_data <- envimp_data %>%
  arrange(datetime)
```


average night temperature 10 pm - 6 am
```{r}
env_night_temp <- envimp_data %>%
  mutate(
    date = as.Date(datetime),                  
    night_date = if_else(hour(datetime) >= 22, date + 1, date)
  ) %>%
  filter(hour(datetime) >= 22 | hour(datetime) < 6) %>%
  group_by(night_date) %>%
  summarise(
    env_temp_night_mean = mean(env_temp, na.rm = TRUE),
    .groups = "drop"
  )

envimp_data <- envimp_data %>%
  mutate(date = as.Date(datetime)) %>%
  left_join(env_night_temp, by = c("date" = "night_date"))
```

calculating daily averages (i.e., 24 hour lags for each participant)
- temperature, PM2.5, PM10, humidity, precipitation, and BV 880
- SD, maximum, minimum, and range for all
- note that the first 23 rows will be empty
```{r}
vars <- c("env_temp","env_PM25","env_PM10","env_rel_humidity", "env_abs_humidity", "env_precip","env_bc_880")

# Create 24-hour rolling averages
envimp_data <- envimp_data %>%
  mutate(
    across(
      all_of(vars),
      list(
        daily_mean = ~ rollapply(.x, 24, mean,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_sd   = ~ rollapply(.x, 24, sd,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_min  = ~ rollapply(.x, 24, min,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_max  = ~ rollapply(.x, 24, max,
                                 align = "right", fill = NA, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )
```

temperature averages "yesterday" (24-48 hours ago)
- again, using rolling averages so first 48 hours will not have data
```{r}
#average temperature
envimp_data <- envimp_data %>%
  mutate(
    env_temp_yesterday =
      lag(
        rollapply(
          env_temp, 24, mean,
          align = "right", fill = NA, na.rm = TRUE
        ),
        24
      )
  )

#SD
envimp_data <- envimp_data %>%
  mutate(
    env_temp_sd_yesterday =
      lag(
        rollapply(
          env_temp_daily_sd, 24, mean,
          align = "right", fill = NA, na.rm = TRUE
        ),
        24
      )
  )
```

temperature averages and SD over the last week
- rolling averages (i.e., last 48 hours, last 72 hours, etc. etc.)
```{r}
#average temperature
envimp_data <- envimp_data %>%
  mutate(
    env_temp_last2d  = rollapply(env_temp,  48, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last3d  = rollapply(env_temp,  72, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last4d  = rollapply(env_temp,  96, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last5d = rollapply(env_temp, 120, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last6d = rollapply(env_temp, 144, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last7d = rollapply(env_temp, 168, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last14d = rollapply(env_temp, 336, mean,
                                 align = "right", fill = NA, na.rm = TRUE)

  )

#SD
envimp_data <- envimp_data %>%
  mutate(
    env_temp_sd_last2d  = rollapply(env_temp_daily_sd,  48, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last3d  = rollapply(env_temp_daily_sd,  72, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last4d  = rollapply(env_temp_daily_sd,  96, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last5d = rollapply(env_temp_daily_sd, 120, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last6d = rollapply(env_temp_daily_sd, 144, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last7d = rollapply(env_temp_daily_sd, 168, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last14d = rollapply(env_temp_daily_sd, 336, mean,
                                 align = "right", fill = NA, na.rm = TRUE)

  )
```

daily average compared to yesterday, past week
```{r}
#average temperature
envimp_data <- envimp_data %>%
  mutate(
    # Daily minus weekly average
    temp_diff_weekly =
      env_temp_daily_mean - env_temp_last7d,

    # Daily minus yesterday average
    temp_diff_yesterday =
      env_temp_daily_mean - env_temp_yesterday
  )

#SD
envimp_data <- envimp_data %>%
  mutate(
    # Daily minus weekly average
    temp_sd_diff_weekly =
      env_temp_daily_sd - env_temp_sd_last7d,

    # Daily minus yesterday average
    temp_sd_diff_yesterday =
      env_temp_daily_sd - env_temp_sd_yesterday
  )
```

hourly temperature lags (not really needed bc of distributed lag model)
```{r}
envimp_data <- envimp_data %>%
  mutate(
    # last 3 hours
    env_temp_last3h = slide_index_dbl(
      .x = env_temp,
      .i = datetime,
      .f = ~ mean(.x, na.rm = TRUE),
      .before = hours(3),
      .complete = FALSE
    ),
    
    # last 7 hours
    env_temp_last7h = slide_index_dbl(
      .x = env_temp,
      .i = datetime,
      .f = ~ mean(.x, na.rm = TRUE),
      .before = hours(7),
      .complete = FALSE
    )
  )
```

add the imputed environmental columns to the dataset
```{r}
#round KORA data to the nearest hour to match with provided environmental data
df$v2ht_starthour <- lubridate::round_date(df$v2ht_starttime, "hour")

#join
merged_data_imp <- df %>%
  left_join(envimp_data, by = c("v2ht_starthour" = "datetime"))
```

```{r}
merged_data_imp$env_temp_night_mean
```


save dataset
```{r}
write.csv(clean_df_env_imp)
saveRDS(merged_data_imp)
```




#6.2. non-imputed environmental data

create a temporary environmental data object
- update the date column and datetime col (use UTC without time jumps to match the variables)
- add main UFP data variables (temperature, Pm2.5, Pm10, black carbon, humidity, precipitation)
```{r}
env_data <- data.frame("date" = as.Date(envraw$date_neu, format = "%d-%B-%Y"))
env_data$hour <- envraw$Hour
env_data$datetime <- as.POSIXct(paste(env_data$date, env_data$hour), 
                                format = "%Y-%m-%d %H", tz = "UTC")

env_data$env_temp <- envraw$Temp_A_LfU_UFP
env_data$env_PM25 <- envraw$PM25_A_LfU_UFP
env_data$env_PM10 <- envraw$PM10_A_LfU_UFP
env_data$env_rel_humidity <- envraw$Feuchte_A_LfU_UFP
env_data$env_precip <- envraw$NI_A_LfU_UFP
env_data$env_bc_880 <- envraw$BC_880_A_LfU_UFP

env_data$env_abs_humidity <-
  (6.112 * exp((17.67 * env_data$env_temp) /
               (envimp_data$env_temp + 243.5)) *
     (envimp_data$env_rel_humidity / 100) * 
     216.7) / 
  (273.15 + envimp_data$env_temp)

# Ensure data are ordered
env_data <- env_data %>%
  arrange(datetime)
```



calculating daily averages (i.e., 24 hour lags for each participant)
- temperature, PM2.5, PM10, humidity, precipitation, and BV 880
- SD, maximum, minimum, and range for all
```{r}
vars <- c("env_temp","env_PM25","env_PM10","env_abs_humidity", "env_rel_humidity","env_precip","env_bc_880")

# Create 24-hour rolling averages
env_data <- env_data %>%
  mutate(
    across(
      all_of(vars),
      list(
        daily_mean = ~ rollapply(.x, 24, mean,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_sd   = ~ rollapply(.x, 24, sd,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_min  = ~ rollapply(.x, 24, min,
                                 align = "right", fill = NA, na.rm = TRUE),
        daily_max  = ~ rollapply(.x, 24, max,
                                 align = "right", fill = NA, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )
```

average night temperature 10 pm - 6 am
```{r}
env_night_temp <- env_data %>%
  mutate(
    date = as.Date(datetime),                  
    night_date = if_else(hour(datetime) >= 22, date + 1, date)
  ) %>%
  filter(hour(datetime) >= 22 | hour(datetime) < 6) %>%
  group_by(night_date) %>%
  summarise(
    env_temp_night_mean = mean(env_temp, na.rm = TRUE),
    .groups = "drop"
  )

env_data <- env_data %>%
  mutate(date = as.Date(datetime)) %>%
  left_join(env_night_temp, by = c("date" = "night_date"))
```


temperature averages yesterday (24-28 hours ago)
```{r}
#average temperature
env_data <- env_data %>%
  mutate(
    env_temp_yesterday =
      lag(
        rollapply(
          env_temp, 24, mean,
          align = "right", fill = NA, na.rm = TRUE
        ),
        24
      )
  )

#SD
env_data <- env_data %>%
  mutate(
    env_temp_sd_yesterday =
      lag(
        rollapply(
          env_temp_daily_sd, 24, mean,
          align = "right", fill = NA, na.rm = TRUE
        ),
        24
      )
  )
```

temperature averages and SD over the last week
```{r}
#average temperature
env_data <- env_data %>%
  mutate(
    env_temp_last2d  = rollapply(env_temp,  48, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last3d  = rollapply(env_temp,  72, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last4d  = rollapply(env_temp,  96, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last5d = rollapply(env_temp, 120, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last6d = rollapply(env_temp, 144, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last7d = rollapply(env_temp, 168, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_last14d = rollapply(env_temp, 336, mean,
                                 align = "right", fill = NA, na.rm = TRUE)

  )

#SD
env_data <- env_data %>%
  mutate(
    env_temp_sd_last2d  = rollapply(env_temp_daily_sd,  48, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last3d  = rollapply(env_temp_daily_sd,  72, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last4d  = rollapply(env_temp_daily_sd,  96, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last5d = rollapply(env_temp_daily_sd, 120, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last6d = rollapply(env_temp_daily_sd, 144, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last7d = rollapply(env_temp_daily_sd, 168, mean,
                                  align = "right", fill = NA, na.rm = TRUE),
    env_temp_sd_last14d = rollapply(env_temp_daily_sd, 336, mean,
                                 align = "right", fill = NA, na.rm = TRUE)

  )
```

daily average compared to yesterday, past week
```{r}
#average temperature
env_data <- env_data %>%
  mutate(
    # Daily minus weekly average
    temp_diff_weekly =
      env_temp_daily_mean - env_temp_last7d,

    # Daily minus yesterday average
    temp_diff_yesterday =
      env_temp_daily_mean - env_temp_yesterday
  )

#SD
env_data <- env_data %>%
  mutate(
    # Daily minus weekly average
    temp_sd_diff_weekly =
      env_temp_daily_sd - env_temp_sd_last7d,

    # Daily minus yesterday average
    temp_sd_diff_yesterday =
      env_temp_daily_sd - env_temp_sd_yesterday
  )
```


hourly temperature lags (not really needed bc of distributed lag model)
```{r}
env_data <- env_data %>%
  mutate(
    # last 3 hours
    env_temp_last3h = slide_index_dbl(
      .x = env_temp,
      .i = datetime,
      .f = ~ mean(.x, na.rm = TRUE),
      .before = hours(3),
      .complete = FALSE
    ),
    
    # last 7 hours
    env_temp_last7h = slide_index_dbl(
      .x = env_temp,
      .i = datetime,
      .f = ~ mean(.x, na.rm = TRUE),
      .before = hours(7),
      .complete = FALSE
    )
  )
```

add the environmental columns to the dataset
```{r}
#round KORA data to the nearest hour to match with provided environmental data
df$v2ht_starthour <- lubridate::round_date(df$v2ht_starttime, "hour")

# drop other time columns to avoid duplicates
env_data <- env_data %>%
  select(-date, -hour)

#join
merged_data <- df %>%
  left_join(env_data, by = c("v2ht_starthour" = "datetime"))
```

#save dataset
```{r}
write.csv(merged_data)
saveRDS(merged_data)

#also save just the non-imputed environmental variables to create CC datasets
saveRDS(env_data)
```





#7. create create complete case datasets for environmental data 
- these datasets have all relevant environmental data for the last 72 hours (originally using 72 prior hours for the crosis-basis)
- start with the environmental df of non-imputed environmental variables

For each environmental variable, create a binary column that notes if there is valid environmental data for each environmental variable for the previous <max_lag> hours. Currently set at 72.
```{r}
max_lag <- 72
env_vars <- c("env_temp", "env_PM25", "env_PM10", "env_abs_humidity", "env_rel_humidity", "env_precip", "env_bc_880")

for (v in env_vars) {
  env_data[[paste0("keep_", v)]] <-
    rollapply(
      env_data[[v]],
      width = max_lag,
      align = "right",
      fill = NA,
      FUN = function(x) all(!is.na(x))
    )
}
```

add the new rows to the df dataset
```{r}
merged_data <- clean_df_w_env %>%
  left_join(
    env_data %>%
      select(datetime, starts_with("keep_")),
    by = c("v2ht_starthour" = "datetime")
  )
```

print a summary of how many visits will be removed
```{r}
merged_data %>%
  select(starts_with("keep_")) %>%
  summarise(across(
    everything(),
    list(
      TRUE_n  = ~ sum(. == TRUE, na.rm = TRUE),
      FALSE_n = ~ sum(. == FALSE, na.rm = TRUE),
      NA_n    = ~ sum(is.na(.)),
      TRUE_pct  = ~ mean(. == TRUE, na.rm = TRUE) * 100,
      FALSE_pct = ~ mean(. == FALSE, na.rm = TRUE) * 100
    )
  )) %>%
  pivot_longer(
    everything(),
    names_to = c("variable", "metric"),
    names_sep = "_(?=[^_]+$)"  # split at last underscore
  ) %>%
  pivot_wider(names_from = metric, values_from = value)
```

check if PM2.5 and PM10 have the same missingness patterns - they are equivalent
```{r}
sum(merged_data$keep_env_PM25 != merged_data$keep_env_PM10, na.rm = TRUE)
```

save dataset with the row indicating missing data
```{r}
saveRDS(merged_data)
```

create datasets for each environmental variable that removes rows without valid data
```{r}
# Per-variable filtered data frames
df_temp     <- merged_data %>% filter(keep_env_temp)
df_pm       <- merged_data %>% filter(keep_env_PM25)
df_rel_humidity <- merged_data %>% filter(keep_env_rel_humidity)
df_abs_humidity <- merged_data %>% filter(keep_env_abs_humidity)
df_precip   <- merged_data %>% filter(keep_env_precip)
df_bc       <- merged_data %>% filter(keep_env_bc_880)

# A strict dataset requiring PM, temp, and humidity variables to be complete
df_all <- merged_data %>% filter(
  keep_env_temp &
  keep_env_PM25 &
  keep_env_rel_humidity
)
```

save
```{r}
outdir <- ""

saveRDS(df_temp,     file.path(outdir, "df_temp.rds"))
saveRDS(df_pm,       file.path(outdir, "df_pm.rds"))
saveRDS(df_rel_humidity, file.path(outdir, "df_rel_humidity.rds"))
saveRDS(df_abs_humidity, file.path(outdir, "df_abs_humidity.rds"))
saveRDS(df_precip,   file.path(outdir, "df_precip.rds"))
saveRDS(df_bc,       file.path(outdir, "df_bc.rds"))
saveRDS(df_all,      file.path(outdir, "df_cc_env.rds"))
```
