
This file represents an exploration of descriptive analyses

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(tidyr)
library(writexl)
library(lubridate)
library(purrr)
library(corrplot)
```

load the rdf file (which includes the factors)
```{r}
clean_df <- readRDS()
envraw <- read.csv()
```


#1. Temperature major statistics
```{r}
clean_df %>%
  summarize(
    n = sum(!is.na(env_temp)),
    mean = mean(env_temp, na.rm = TRUE),
    sd = sd(env_temp, na.rm = TRUE),
    var = var(env_temp, na.rm = TRUE),
    min = min(env_temp, na.rm = TRUE),
    q25 = quantile(env_temp, 0.25, na.rm = TRUE),
    median = median(env_temp, na.rm = TRUE),
    q75 = quantile(env_temp, 0.75, na.rm = TRUE),
    max = max(env_temp, na.rm = TRUE),
    range = max - min,
    IQR = IQR(env_temp, na.rm = TRUE)
  )
``` 

#2. correlation for environmental variables
```{r}
env_data <- data.frame("date" = as.Date(envraw$date_neu, format = "%d-%B-%Y"))
env_data$hour <- envraw$Hour
env_data$datetime <- as.POSIXct(paste(env_data$date, env_data$hour), 
                                format = "%Y-%m-%d %H", tz = "UTC")
```

add UFP data variables
- temperature, Pm2.5, Pm10, black carbon, humidity, precipitation
```{r}
env_data$env_temp_ufp <- envraw$Temp_A_LfU_UFP
env_data$env_PM25_ufp <- envraw$PM25_A_LfU_UFP
env_data$env_PM10_ufp <- envraw$PM10_A_LfU_UFP
env_data$env_humidity_ufp <- envraw$Feuchte_A_LfU_UFP
env_data$env_precip_ufp <- envraw$NI_A_LfU_UFP
env_data$env_bc_880_ufp <- envraw$BC_880_A_LfU_UFP
```

```{r}
numeric_df <- env_data %>% 
  select(where(is.numeric)) %>%
  select(-hour) %>%
  select(starts_with("env_temp"), everything())

clean_names <- gsub("_ufp", "", names(numeric_df))
clean_names <- gsub("env_", "", clean_names)
clean_names <- gsub("_avg", "", clean_names)
colnames(numeric_df) <- clean_names

cor_mat <- cor(numeric_df, use = "pairwise.complete.obs", method = "pearson")

corrplot(cor_mat, method = "color", type = "upper",
         tl.col = "black", 
         tl.cex = 0.8, 
         addCoef.col = "black",
         number.cex = 0.6)
```


#3. calculate the number of visits for each participant
remove visits with missing spirometry data
```{r}
visit_summary <- clean_df %>%
  filter(!is.na(v2ht_starttime)) %>%
  group_by(ID) %>%
  summarise(num_visits = n(), .groups = "drop") %>%
  count(num_visits, name = "num_participants") %>%
  mutate(percent = num_participants / sum(num_participants) * 100)
```

#4. Explore covariates and outcomes

separate datasets based on variables -- column layouts differ
```{r}
#medication df
df_med <- clean_df %>%
  select(ID, v2x_visit, v2i_panel, v2h_examdate, starts_with("ATC"))

#spirometry df
df_spirometry <- clean_df %>%
  select(ID, v2x_visit, v2i_panel, starts_with("v2h"))

#covariates
df_covariates <- clean_df %>%
  select(
    -starts_with("ATC", ignore.case = TRUE),
    -starts_with("v2h", ignore.case = TRUE)
  )
```


#4.1 start with covariates dataset

group based on participant id (one row per participant)
- Get the baseline value for all characteristics. Height and weight recorded every time. All other variables were only recorded at the first visit.
```{r}
df <- df_covariates %>%
  group_by(ID) %>%
  filter(v2x_visit == 1) %>%
  ungroup()
```


functions to create plots and table 1

create plots
```{r}
make_plots <- function(df, meta, bin_width) {
  
  plot_list <- list()
  
  for (v in names(meta)) {
    
    label  <- meta[[v]]$label
    values <- meta[[v]]$values
    
    # categorical variables
    if (!is.null(values)) {
      
      p <- ggplot(df, aes(x = factor(.data[[v]]))) +
        geom_bar(fill = "chartreuse4") +
        scale_x_discrete(labels = values) +
        labs(
          title = label,
          x = label,
          y = "Count"
        ) +
        theme_minimal() +
        theme(
          plot.title  = element_text(size = 10, face = "bold"),
          axis.title.y = element_text(size = 5),
          axis.title.x = element_blank(),
          axis.text    = element_text(size = 6)
        )
      
    } else {
      # continuous variables
      p <- ggplot(df, aes(x = .data[[v]])) +
        geom_histogram(
          fill = "chartreuse4",
          binwidth = bin_width[[v]]
        ) +
        labs(
          title = label,
          x = label,
          y = "Count"
        ) +
        theme_minimal() +
        theme(
          plot.title  = element_text(size = 10, face = "bold"),
          axis.title.y = element_text(size = 5),
          axis.title.x = element_blank(),
          axis.text    = element_text(size = 5)
        )
    }
    
    plot_list[[v]] <- p
  }
  
  return(plot_list)
}
```

create table 1
```{r}
make_table1 <- function(df, meta) {
  
  table1_list <- list()
  
  for (v in names(meta)) {
    
    var_info <- meta[[v]]
    
    #categorical variables
    if (!is.null(var_info$values)) {
      
      tab <- df %>%
        mutate(
          val = factor(
            .data[[v]],
            levels = names(var_info$values),
            labels = var_info$values
          )
        ) %>%
        
        group_by(v2i_panel, val) %>%
        summarise(n = n(), .groups = "drop") %>%
        
        pivot_wider(
          names_from = v2i_panel,
          values_from = n,
          values_fill = 0
        ) %>%
        
        mutate(All = rowSums(across(-val))) %>%
        
        mutate(across(
          -c(val, All),
          ~ paste0(.x, " (", round(.x / sum(.x) * 100, 1), "%)")
        )) %>%
        
        mutate(All = paste0(All, " (", round(All / sum(All) * 100, 1), "%)")) %>%
        
        mutate(variable = var_info$label, category = val) %>%
        select(variable, category, All, everything(), -val)
      
      # Fisher's exact test
      chi_tab <- table(df[[v]], df$v2i_panel)
      pval <- tryCatch(
        fisher.test(chi_tab)$p.value,
        error = function(e) NA
      )
      
      tab <- tab %>%
        mutate(`p-value` = round(pval, 3))
      
      table1_list[[v]] <- tab
      
    } else {
      
      #continuous variables
      tab <- df %>%
        group_by(v2i_panel) %>%
        summarise(
          mean_val = mean(.data[[v]], na.rm = TRUE),
          sd_val   = sd(.data[[v]], na.rm = TRUE),
          .groups = "drop"
        ) %>%
        
        mutate(mean_sd = paste0(
          round(mean_val, 1), " (", round(sd_val, 1), ")"
        )) %>%
        
        select(v2i_panel, mean_sd) %>%
        pivot_wider(names_from = v2i_panel, values_from = mean_sd) %>%
        
        mutate(All = paste0(
          round(mean(df[[v]], na.rm = TRUE), 1),
          " (",
          round(sd(df[[v]], na.rm = TRUE), 1),
          ")"
        )) %>%
        
        mutate(variable = var_info$label, category = "") %>%
        select(variable, category, All, everything())
      
      # ANOVA
      aov_model <- aov(df[[v]] ~ df$v2i_panel)
      pval <- summary(aov_model)[[1]][["Pr(>F)"]][1]
      
      tab <- tab %>%
        mutate(`p-value` = round(pval, 3))
      
      table1_list[[v]] <- tab
    }
  }
  
  # Combine all variables into one table
  bind_rows(table1_list)
}
```


#4.1.1 demographics
```{r}
#categorical
demographics_meta <- list(
  v2i_panel = list(
    label = "Panel",
    values = c(
      "COPD" = "COPD",
      "Hypertonie" = "Hypertension",
      "T2DM" = "Type 2\n Diabetes")),
  
  v2it_age = list(
    label  = "Age"),
  
  v2it_sex = list(
    label  = "Sex", 
    values = c("1" = "Male", "2" = "Female")), #sex transformed and sex are equivalent

  v2it_borngermany = list(
    label  = "Born in Germany",
    values = c("1" = "Yes", "0" = "No")),

  v2i_marital = list(
    label  = "Marital Status",
    values = c(
      "1" = "Married",
      "2" = "Single",
      "3" = "Divorced",
      "4" = "Widowed")),

  pt_eduqual = list(
    label  = "Highest Vocational\nQualification",
    values = c(
      "1" = "No\nqualification",
      "2" = "Vocational\nschool/\nApprenticeship",
      "3" = "Technician/\nMaster\nschool",
      "4" = "Engineering school/\nPolytechnic",
      "5" = "Technical\ncollege/\nUniversity")),

  pt_eduyears = list(
    label = "Years of Schooling"),
  
  v2it_employment = list(
    label  = "Employment",
    values = c(
      "1"    = "Full-time",
      "2"    = "Part-time",
      "3"    = "Retired",
      "4"    = "Unemp-\nloyed")),
  
  v2it_equivalizedincomecategories = list(
    label = "Equivalized Household\n Income Compared\nto Median",
     values = c(
      "1"  = "Under\n50%",
      "2"  = "Under\n100%",
      "3"  = "Under\n150%",
      "4"  = "Under\n200%",
      "5"  = "over\n200%"))
)
  
#continuous
demographics_binwidth <- c(
  pt_eduyears = 1,
  v2it_equivalizedincome = 300,
  v2it_age = 3
)
```

```{r}
demographics_table1 <- make_table1(df, demographics_meta)

plots <- make_plots(df, demographics_meta, demographics_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 3)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```


#4.1.2 health behaviors
```{r}
# Categorical variables metadata
healthbehaviors_meta <- list(
  v2it_livepartner = list(
    label  = "Live with Partner",
    values = c("1" = "Yes", "0" = "No")),
  
  v2it_smoking = list(
    label  = "Smoking Status",
    values = c(
      "1" = "current\nsmoker", 
      "2" = "previous\nsmoker",
      "0" = "never/smoker")),
  
  v2i_numcigarettes = list(
    label  = "Number of Cigarettes\nper Day"),
  
  v2it_othertobacco = list(
    label  = "Smoking Status\n(Other Tobacco Products)",
    values = c("1" = "Yes", "0" = "No")),
  
  v2it_secondhandsmoke = list(
    label  = "Secondhand Smoke Exposure",
    values = c("1" = "Yes", "0" = "No")),
  
  v2i_wintersports = list(
    label  = "Sports in Winter",
    values = c(
      "1" = ">2\nhs/week",
      "2" = "1-2\nhs/week",
      "3" = "<1 hour\n/week",
      "4" = "No activity")),
  
  v2i_summersports = list(
    label  = "Sports in Summer",
    values = c(
      "1" = ">2\nhs/week",
      "2" = "1-2\nhs/week",
      "3" = "<1 hour\n/week",
      "4" = "No activity"))
)

# Continuous/numeric variables bin width
healthbehaviors_binwidth <- list(
  v2i_numcigarettes = 3
)
```

```{r}
healthbehaviors_table1 <- make_table1(df, healthbehaviors_meta)

plots <- make_plots(df, healthbehaviors_meta, healthbehaviors_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 3)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```


#4.1.3 self-reported exposures
```{r}
exposures_meta <- list(
  v2i_homeheating = list(
    label  = "Home Heating Method",
    values = c(
      "1" = "Central/external\nheat source",
      "2" = "Single stove or\nburner inside dwelling")),
  v2it_gascooking = list(
    label  = "Gas Stove Used\nfor Cooking",
    values = c("1" = "Yes", "0" = "No")),
  v2it_coolhome = list(
    label  = "Living Area Can Be Kept\nCool on Hot Days",
    values = c("1" = "Yes", "0" = "No")),
  v2it_ac = list(
    label  = "Air Conditioning at Home",
    values = c("1" = "Yes", "0" = "No")),
  v2i_sunwork = list(
    label  = "Exposure to Direct\nSunlight at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_directcoldwork = list(
    label  = "Exposure to Direct\nCold at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_acwork = list(
    label  = "Air Conditioning\nat Workplace",
    values = c("1" = "Yes", "0" = "No")),
  v2i_heatwork = list(
    label  = "Exposure to Technical\nHeat Sources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_coldwork = list(
    label  = "Exposure to Technical\nCold Sources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_moisturework = list(
    label  = "Exposure to Moisture\nSources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_uniformwork = list(
    label  = "Prescribed Work Clothing\n(Uniform / Protective Clothing)",
    values = c("1" = "Yes", "0" = "No"))
)

exposures_binwidth <- list()
```

```{r}
exposures_table1 <- make_table1(df, exposures_meta)

plots <- make_plots(df, exposures_meta, exposures_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 4)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```


#4.1.4 health status
```{r}
# Categorical variables metadata
healthstatus_meta <- list(
  v2i_selfhealth = list(
    label  = "Self-rated Health",
    values = c(
      "1" = "Very good",
      "2" = "Good",
      "3" = "Less good",
      "4" = "Bad")),
  v2i_selfhealthcompared = list(
    label  = "Self-rated Health Compared\nto Same Age Peers",
    values = c(
      "1" = "Better",
      "3" = "No change",
      "2" = "Worse",
      "4" = "don't\nknow")),
  v2it_diabetes = list(
    label  = "Diabetes",
    values = c("1" = "Yes", "0" = "No")),
  v2it_hypertension = list(
    label  = "High Blood Pressure",
    values = c("1" = "Yes", "0" = "No")),
  v2it_sleepapnea = list(
    label  = "Sleep Apnea",
    values = c("1" = "Yes", "0" = "No")),
  v2it_hayfever = list(
    label  = "Hay Fever",
    values = c("1" = "Yes", "0" = "No")),
  v2it_asthma = list(
    label  = "Asthma",
    values = c("1" = "Yes,\ncurrent", "2" = "Yes,\n cured", "0" = "No")),
  v2it_coughsymptoms = list(
    label  = "Cough Symptoms",
    values = c("1" = "Yes", "0" = "No")),
  v2it_sputumsymptoms = list(
    label  = "Sputum Symptoms",
    values = c("1" = "Yes", "0" = "No")),
  v2it_copd = list(
    label  = "COPD",
    values = c("1" = "Yes", "0" = "No")),
  v2it_emphysema = list(
    label  = "Emphysema",
    values = c("1" = "Yes", "0" = "No"))
)

healthstatus_binwidths <- list()
```

```{r}
healthstatus_table1 <- make_table1(df, healthstatus_meta)

plots <- make_plots(df, healthstatus_meta, healthstatus_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 4)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```


#4.1.5 anthropometry
```{r}
anthropometry_meta <- list(
  v2it_height = list(
    label = "Height (cm)"),
  v2it_weight = list(
    label = "Weight (kg)"),
  v2it_waist = list(
    label = "Waist Circumference (cm)"),
  v2it_hip = list(
    label = "Hip Circumference (cm)"),
  v2i_bmi = list(
    label = "BMI"),
  v2it_bmiclass = list(
    label = "BMI",
    values = c(
      "0" = "Underweight",
      "1" = "Normal",
      "2" = "Overweight",
      "3" = "Obese")),
  v2it_whr = list(
    label = "Waist Hip Ratio"),
  v2it_whrclass = list(
    label = "Waist-Hip Ratio",
    values = c(
      "0" = "Low risk",
      "1" = "High risk"))
)

# Continuous / numeric variables bin width
anthropometry_binwidth <- list(
  v2it_height = 2,
  v2it_weight = 3,
  v2it_waist  = 3,
  v2it_hip    = 3,
  v2i_bmi     = 1,
  v2it_whr    = 0.05
)
```

```{r}
anthropometry_table1 <- make_table1(df, anthropometry_meta)

plots <- make_plots(df, anthropometry_meta, anthropometry_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 4)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```




#5. medication
A10   Diabetes
C02   Direct antihypertensives
C03   Diuretics
C07 	Beta-blockers
C08 	Calcium channel blockers
C09 	ACE inhibitors & ARBs
S01G	Decongestants and anti-allergens
R01A	Decongestants and othe tropical rhinologicals
R03   Asthma, COPD, Bronchodilators
R06   Allergies (antihistamines)

get the number of participants recording medication

collapse on each participant ID, only include those select variables
```{r}
df_med_small <- df_med %>%
  group_by(ID) %>%
  summarise(
    v2i_panel = first(v2i_panel),
    across(
      c(ATC_R03, ATC_R06, ATC_S01G, ATC_R01A, ATC_A10, ATC_C02, ATC_C03, ATC_C07, ATC_C08, ATC_C09),
      ~ as.integer(any(.x == 1, na.rm = TRUE))
    )
  )
```

get the counts and percent of each arm that take the medication 
```{r}
medications <- c("ATC_R03","ATC_R06","ATC_S01G","ATC_R01A","ATC_A10",
                 "ATC_C02","ATC_C03","ATC_C07","ATC_C08","ATC_C09")

table_list <- list()

for (med in medications){

  # counts
  table <- df_med_small %>%
    group_by(v2i_panel, !!sym(med)) %>% 
    summarize(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = v2i_panel, values_from = n, values_fill = 0)

  # fisher test
  chi_tab <- table %>% select(-all_of(med))

  pval <- tryCatch(
    fisher.test(chi_tab)$p.value,
    error = function(e) NA
  )

  # add percentages + clean
  table <- table %>%
    mutate(All = rowSums(across(-all_of(med)))) %>%
    mutate(across(-all_of(med),
                  ~ paste0(.x, " (", round(.x / sum(.x) * 100, 1), "%)"))) %>%
    ungroup() %>%
    mutate(
      variable = med,
      'p-value' = round(pval, 3)
    ) %>%
    relocate(variable) %>%
    relocate(All, .after = all_of(med)) %>%
    rename(category = all_of(med))

  table_list[[length(table_list) + 1]] <- table
}

medications_table1 <- bind_rows(table_list)
```

get the number of medications each participant takes
(note that this isn't ideal bc it doesn't have all potential medications)
```{r}
total_meds <- df_med_small %>%
  mutate(
    total_meds = rowSums(across(c(ATC_R03, ATC_R06, ATC_S01G, ATC_R01A, ATC_A10, 
                                  ATC_C02, ATC_C03, ATC_C07, ATC_C08, ATC_C09)))) %>%
  select (total_meds, ID, v2i_panel) %>%
  count(v2i_panel, total_meds) %>%
  tidyr::pivot_wider(
    names_from = total_meds,
    values_from = n,
    values_fill = 0
  )

total_meds
```


#6. environmental exposures
 
```{r}
#categorical
environmentalexposures_meta <- list(
  env_temp_ufp_daily = list(
    label  = "Daily Temperature"),
  env_PM25_ufp_daily = list(
    label  = "Daily PM2.5"),
  env_PM10_ufp_daily = list(
    label  = "Daily PM10"),
  env_humidity_ufp_daily = list(
    label  = "Daily Humidity"),
  env_bc_880_ufp_daily = list(
    label  = "Daily Black Carbon")
)
  
#continuous
environmentalexposures_binwidth <- c(
  env_temp_ufp_daily = 1,
  env_PM25_ufp_daily = 1,
  env_PM10_ufp_daily = 2,
  env_humidity_ufp_daily = 3,
  env_bc_880_ufp_daily = 50
)
```

```{r}
environmentalexposures_table1 <- make_table1(df, environmentalexposures_meta)

plots <- make_plots(df, environmentalexposures_meta, environmentalexposures_binwidth)

combined_plot <- wrap_plots(plots, ncol = 3, nrow = 4)
combined_plot

ggsave("", combined_plot,
       width = 7, height = 5, dpi = 600)
```



#7. Combine all variables into one larger Table 1
```{r}
table1 <- rbind(demographics_table1,
                healthbehaviors_table1,
                healthstatus_table1,
                anthropometry_table1,
                exposures_table1,
                medications_table1,
                environmentalexposures_table1)

write_xlsx(table1, path = "")
```



#8. look at the variables that change over time (excluding pulmonary function)
anthropometric: height, weight
medications: "R03","R06","S01G","R01A","A10","C02","C03","C07","C08","C09"
id, visit number, date
      
weight changes over time
```{r}
change_df <- clean_df %>%
  group_by(ID) %>%
  summarise(
    # baseline & final (last observed visit)
    baseline_weight = v2it_weight[v2x_visit == 1][1],
    final_weight    = v2it_weight[v2x_visit == max(v2x_visit, na.rm = TRUE)][1],
    last_visit      = max(v2x_visit, na.rm = TRUE),

    # change scores
    weight_change = final_weight - baseline_weight,
    # weight descriptives across visits
    weight_min   = min(v2it_weight, na.rm = TRUE),
    weight_max   = max(v2it_weight, na.rm = TRUE),
    weight_range = weight_max - weight_min,
    weight_sd    = sd(v2it_weight, na.rm = TRUE),

    .groups = "drop"
  )
```

plot weight changes over time
```{r}
stats <- c("weight")

# Create empty list to store the plots
plots <- list()

# Loop through stats (weight)
for(stat in stats) {
  # create 3 plots
  plots[[paste0(stat, "_range")]] <- ggplot(change_df, aes_string(x = paste0(stat, "_range"))) +
    geom_histogram(bins = 20, fill = "steelblue", color = "black") +
    labs(title = paste0("Range of ", stat), x = paste0("Range of ", stat), y = "Count") +
    theme_minimal()

  plots[[paste0(stat, "_change")]] <- ggplot(change_df, aes_string(x = paste0(stat, "_change"))) +
    geom_histogram(bins = 20, fill = "steelblue", color = "black") +
    labs(title = paste0("Change in ", stat), x = paste0("Change in ", stat), y = "Count") +
    theme_minimal()

  plots[[paste0(stat, "_sd")]] <- ggplot(change_df, aes_string(x = paste0(stat, "_sd"))) +
    geom_histogram(bins = 20, fill = "steelblue", color = "black") +
    labs(title = paste0("SD of ", stat), x = paste0("SD of ", stat), y = "Count") +
    theme_minimal()
}

# Combine all plots using patchwork
combined_plot <- (plots$weight_range | plots$weight_change | plots$weight_sd)

# Display the combined plot
combined_plot

```

look at medication use across seasons and months
```{r}
# Example of medication columns
medications <- c("ATC_R03","ATC_R06","ATC_S01G","ATC_R01A","ATC_A10",
                 "ATC_C02","ATC_C03","ATC_C07","ATC_C08","ATC_C09")

# Create a new column for the season, month
clean_df <- clean_df %>%
  mutate(
    v2h_examdate = as.POSIXct(v2h_examdate),
    season = case_when(
      month(v2h_examdate) %in% c(12, 1, 2) ~ "Winter",
      month(v2h_examdate) %in% c(3, 4, 5) ~ "Spring",
      month(v2h_examdate) %in% c(6, 7, 8) ~ "Summer",
      month(v2h_examdate) %in% c(9, 10, 11) ~ "Fall",
      TRUE ~ NA_character_
    ),
    month = lubridate::floor_date(v2h_examdate, "month")
  )

# Summarize the number of medications used per season
medication_summary <- clean_df %>%
  group_by(season) %>%
  summarise(across(all_of(medications), ~ sum(. == 1, na.rm = TRUE), .names = "med_{.col}"),
            .groups = "drop")
print(medication_summary)

# Summarize monthly counts for each medication
monthly_summary <- clean_df %>%
  group_by(month) %>%
  summarise(across(all_of(medications),
                   ~ sum(. == 1, na.rm = TRUE),
                   .names = "{.col}"),
            .groups = "drop") %>%
  pivot_longer(cols = all_of(medications),
               names_to = "medication",
               values_to = "count")

# Time series plot (faceted)
ggplot(monthly_summary, aes(x = month, y = count, color = medication)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ medication, scales = "free_y") +
  labs(title = "Monthly Medication Usage Over Time",
       x = "Month",
       y = "Count of Usage") +
  theme_minimal() +
  theme(legend.position = "none")

```






#9. Look at the distribution across three arms
continuous
```{r}
make_histogram_plots <- function(df, var, var_name) {
  var <- sym(var)   # convert string to symbol
  
  # create one plot per subject type
  plots <- df %>%
    split(.$v2i_panel) %>%
    lapply(function(d) {
      ggplot(d, aes(x = !!var)) +
        geom_histogram(
          fill = "steelblue",
          binwidth = 3
        ) +
        scale_y_continuous(limits = c(0, 11)) +
        labs(
          title = paste(var_name, " —", unique(d$v2i_panel)),
          x = var_name,
          y = "Count"
        ) +
        theme_minimal() +
        theme(
          plot.title  = element_text(size = 10, face = "bold"),
          axis.title.y = element_text(size = 5),
          axis.title.x = element_blank(),
          axis.text    = element_text(size = 5)
        )
    })
  
  # arrange in a 3×3 grid
  p <- wrap_plots(plots, ncol = 3, nrow = 3)
  return(p)
}
```

categorical
```{r}
make_categorical_plots <- function(df, var, var_name, labels) {
  var <- sym(var)

  plots <- df %>%
    split(.$v2i_panel) %>%
    lapply(function(d) {

      # Ensure all categories appear in every subset
      d <- d %>% mutate(
        !!var := factor(as.character(!!var), levels = names(labels))
      )

      ggplot(d, aes(x = !!var)) +
        geom_bar(fill = "steelblue") +
        scale_x_discrete(labels = labels, drop = FALSE) +
        scale_y_continuous(limits = c(0, 27)) +
        labs(
          title = paste(var_name, "—", unique(d$v2i_panel)),
          x = var_name,
          y = "Count"
        ) +
        theme_minimal() +
        theme(
          plot.title  = element_text(size = 10, face = "bold"),
          axis.title.y = element_text(size = 5),
          axis.title.x = element_blank(),
          axis.text    = element_text(size = 5)
        )
    })

  wrap_plots(plots, ncol = 3, nrow = 3)
}
```


age
```{r}
p_age <- make_histogram_plots(df, "v2it_age", "Age")
p_age
```

sex
```{r}
p_sex <- make_categorical_plots(
  df,
  var = "v2it_sex",
  var_name = "Sex",
  labels = c("1" = "Male", "2" = "Female")
)

p_sex
```

income
```{r}
p_income <- make_categorical_plots(
  df,
  var = "v2it_equivalizedincomecategories",
  var_name = "Equivalized Income",
  labels = c("1" = "50%", "2" = "100%", "3" = "150%", "4" = "200%", "5" = "over\n200%")
)

p_income
```

smoking
```{r}
p_smoking <- make_categorical_plots(
  df,
  var = "v2it_smoking",
  var_name = "Smoking",
  labels = c("1" = "current\nsmoker", "2" = "previous\nsmoker", "0" = "never\nsmoker")
)

p_smoking
```

BMI (with dashed lines for overweight/obesity)
```{r}
# create one plot per subject type
plots <- df %>%
  split(.$v2i_panel) %>%
  lapply(function(d) {
    ggplot(d, aes(x = v2i_bmi)) +
      geom_histogram(
        fill = "steelblue",
        binwidth = 2
      ) +
      geom_vline(xintercept = 25, color = "red", linetype = "dashed", size = 0.6) +
      geom_vline(xintercept = 30, color = "red", linetype = "dashed", size = 0.6) +
      scale_y_continuous(limits = c(0, 11)) +
      scale_x_continuous(limits = c(17, 47)) +
      labs(
        title = paste("BMI — ", unique(d$v2i_panel)),
        x = "BMI",
        y = "Count"
      ) +
      theme_minimal() +
      theme(
        plot.title  = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 5),
        axis.title.x = element_blank(),
        axis.text    = element_text(size = 5)
      )
  })

# arrange them in a 3×3 grid
p_bmi <- wrap_plots(plots, ncol = 3, nrow = 3)

p_bmi
```

whr, also stratified by gender (with dashed lines for abdominal obesity)
```{r}
plots_whr <- df %>%
  split(list(.$v2i_panel, .$v2it_sex), drop = TRUE) %>%
  lapply(function(d) {

    # choose threshold by sex
    thr <- ifelse(unique(d$v2it_sex) == "2", 0.85, 1.0)
    sex_label <- ifelse(unique(d$v2it_sex) == 1, "male", "female")

    ggplot(d, aes(x = v2it_whr)) +
      geom_histogram(fill = "darkorange", binwidth = 0.02) +
      geom_vline(xintercept = thr, color = "red", linetype = "dashed", size = 0.6) +
      scale_y_continuous(limits = c(0, 6)) +
      coord_cartesian(xlim = c(0.73, 1.17)) +
      labs(
        title = paste("WHR — ", unique(d$v2i_panel), "/", sex_label),
        x = "Waist–Hip Ratio",
        y = "Count"
      ) +
      theme_minimal() +
      theme(
        plot.title  = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 5),
        axis.title.x = element_blank(),
        axis.text    = element_text(size = 5)
      )
  })

p_whr <- wrap_plots(plots_whr, ncol = 3, nrow = 4)
p_whr
```


combing the occupational exposures into one graph (restricting to those who work)
```{r}
workexposures_meta <- list(
  v2i_sunwork = list(
    label  = "Exposure to Direct\nSunlight at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_directcoldwork = list(
    label  = "Exposure to Direct\nCold at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_acwork = list(
    label  = "Air Conditioning\nat Workplace",
    values = c("1" = "Yes", "0" = "No")),
  v2i_heatwork = list(
    label  = "Exposure to Technical\nHeat Sources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_coldwork = list(
    label  = "Exposure to Technical\nCold Sources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_moisturework = list(
    label  = "Exposure to Moisture\nSources at Work",
    values = c("1" = "Yes", "0" = "No")),
  v2i_uniformwork = list(
    label  = "Prescribed Work Clothing\n(Uniform / Protective Clothing)",
    values = c("1" = "Yes", "0" = "No"))
)
```

```{r}
vars <- names(workexposures_meta)

# Recode values using meta list
df_long <- df %>%
  filter(v2i_employmenttype != 4) %>% # Filter out those that are not employed
  mutate(across(all_of(vars), ~ workexposures_meta[[cur_column()]]$values[as.character(.)])) %>%
  pivot_longer(cols = all_of(vars), names_to = "variable", values_to = "response")

# Count only "Yes"
yes_counts <- df_long %>%
  filter(response == "Yes") %>%
  group_by(variable) %>%
  summarise(count_yes = n()) %>%
  ungroup() %>%
  mutate(label = map_chr(variable, ~ workexposures_meta[[.x]]$label))

plot_list <- list()

p <- ggplot(yes_counts, aes(x = reorder(label, count_yes), y = count_yes)) +
      geom_bar(stat = "identity", fill = "chartreuse4") +
      labs(
        title = "Occupational Exposures",
        x = "Exposures",
        y = "Count"
      ) +
      theme_minimal() +
      theme(
        plot.title  = element_text(size = 13.5, face = "bold"),
        axis.title.y = element_text(size = 5),
        axis.title.x = element_blank(),
        axis.text    = element_text(size = 5)
      )

plot_list[[1]] <- p

combined_plot <- wrap_plots(plot_list, ncol = 1, nrow = 3)
combined_plot
```







