
This file contains code to explore general trends between temperature and lung function health outcomes, including DLNM models, GEEs, and GAMs.

```{r}
library(dplyr)
library(tidyr)
library(dlnm)
library(lme4)
library(splines)
library(mgcv)
library(ggplot2)
library(patchwork)
library(geepack)
library(corrplot)
```

get datasets
```{r}
outdir <- ""

envraw <- read.csv(file.path(outdir, "exposure data", "LFU_Rest_Augsburg_2022_25.csv"))
imp_env <- readRDS(file.path(outdir, "env_imputed.rds"))
df <- readRDS(file.path(outdir, "clean_df_w_env_imp.rds"))

```


setting up variables
```{r}
# Covariates (month is already included as a fixed effect or spline term)
cov_fixed <- c("v2it_sex", "v2it_age", "v2i_panel", "v2it_smoking", "v2ht_respinfection")
cov_fixed <- c("v2ht_respinfection")
id_var <- "ID"

outcomes <- list(
  FEV1FVC = list(
    var   = "v2h_FEV1FVC",
    title = "FEV1/FVC (%)"
  ),
  FEV1 = list(
    var   = "v2h_FEV1_mL",
    title = "FEV1 (mL)"
  ),
  FVC = list(
    var   = "v2h_FVC_mL",
    title = "FVC (mL)"
  )
)
```

labels
```{r}
nice_env_labels <- function(var) {
  labels <- c(
    "env_temp_daily_mean" = "Temperature Daily Avg",
    "env_temp_daily_sd"  = "Temperature SD",
    "env_temp_daily_min" = "Temperature Min",
    "env_temp_daily_max" = "Temperature Max",
    "env_PM25_daily_mean" = "PM2.5 Daily Avg",
    "log_env_PM25_daily_mean" = "log PM2.5 Daily Avg",
    "env_PM10_daily_mean" = "PM10 Daily Avg",
    "log_env_PM10_daily_mean" = "log PM10 Daily Avg",
    "env_bc_880_daily_mean" = "BC 880 Daily Avg",
    "log_env_bc_880_daily_mean" = "log BC 880 Daily Avg",
    "env_rel_humidity_daily_mean" = "Reltative Humidity Daily Avg",
    "env_abs_humidity_daily_mean" = "Absolute Humidity Daily Avg",
    "env_precip_daily_mean" = "Precipitation Daily Avg",
    "log_env_precip_daily_mean" = "log Precipitation Daily Avg",
    "env_temp_last3h" = "Temperature 3 hours prior",
    "env_temp_last7h" = "Temperature 7 hours prior",
    "env_temp_yesterday" = "Yesterday Temperature Avg",
    "env_temp_last2d" = "2-Day Temperature Avg",
    "env_temp_last3d" = "3-Day Temperature Avg",
    "env_temp_last4d" = "4-Day Temperature Avg",
    "env_temp_last5d" = "5-Day Temperature Avg",
    "env_temp_last6d" = "6-Day Temperature Avg",
    "env_temp_last7d" = "Weekly Temperature Avg",
    "env_temp_last14d" = "2 Week Temperature Avg",
    "temp_diff_yesterday" = "Today - Yesterday\nTemperature Average",
    "temp_diff_weekly" = "Today - Weekly\nTemperature Average",
    "env_temp_night_mean" = "Temperature Night Average"
  )
  
  var <- as.character(var)
  
  out <- labels[var]
  
  out[is.na(out)] <- var[is.na(out)]
  
  return(out)
}
```

creating additional variables
```{r}
#season
df$season <- with(
  df,
  factor(
    dplyr::case_when(
      lubridate::month(v2h_examdate) %in% c(12, 1, 2) ~ "Winter",
      lubridate::month(v2h_examdate) %in% c(3, 4, 5)  ~ "Spring",
      lubridate::month(v2h_examdate) %in% c(6, 7, 8)  ~ "Summer",
      lubridate::month(v2h_examdate) %in% c(9, 10, 11) ~ "Fall",
      TRUE ~ NA_character_
    ),
    levels = c("Winter", "Spring", "Summer", "Fall")
  )
)

df$month_num <- as.numeric(as.character(df$month))

#hour of day

df <- df %>%
  mutate(
    hourofday = lubridate::hour(v2ht_starttime),
    dayofweek = lubridate::wday(v2ht_starttime, label = TRUE, abbr = FALSE),
    weekofyear = lubridate::week(v2ht_starttime)
  )

```

helper method so that plots are outputted in a helpful format
```{r}
paginate_exposure_plots <- function(
    plot_list,
    outcomes,
    env_vars,
    n_cols = 4   # env_vars per page (columns)
) {
  library(patchwork)
  
  n_out <- length(outcomes)
  n_env <- length(env_vars)
  
  if (length(plot_list) != n_out * n_env) {
    stop("plot_list length must equal outcomes × env_vars")
  }
  
  # helper: index into your outcome-major list
  get_plot <- function(out_j, env_i) {
    (out_j - 1) * n_env + env_i
  }
  
  # paginate env_vars
  col_groups <- split(seq_len(n_env),
                      ceiling(seq_len(n_env) / n_cols))
  
  for (i in seq_along(col_groups)) {
    cols_idx <- col_groups[[i]]
    
    # build rows = outcomes
    page_rows <- lapply(seq_len(n_out), function(out_j) {
      
      row_plots <- lapply(cols_idx, function(env_i) {
        plot_list[[ get_plot(out_j, env_i) ]]
      })
      
      wrap_plots(row_plots, ncol = length(cols_idx))
    })
    
    # stack outcome rows vertically
    page <- wrap_plots(page_rows, ncol = 1) +
      plot_annotation(title = paste("Page", i))
    
    print(page)
  }
}
```




#1. visualizations of exposures

variables
```{r}
env_vars <- c("env_temp_last3h","env_temp_daily_mean","env_temp_last7d","env_temp_daily_sd","env_temp_daily_min","env_temp_daily_max","env_abs_humidity_daily_mean","env_rel_humidity_daily_mean","env_PM25_daily_mean","env_PM10_daily_mean","env_bc_880_daily_mean","env_precip_daily_mean")
```

transformations (log, etc.)
```{r}
df <- df %>%
  mutate(
    log_env_bc_880_daily_mean   = log(env_bc_880_daily_mean + 1),
    log_env_precip_daily_mean   = log(env_precip_daily_mean + 1),
    log_env_PM25_daily_mean     = log(env_PM25_daily_mean + 1),
    log_env_PM10_daily_mean     = log(env_PM10_daily_mean + 1),
    
    env_precip_daily_mean_p99 = ifelse(
      env_precip_daily_mean <= quantile(env_precip_daily_mean, 0.99, na.rm = TRUE),
      env_precip_daily_mean,
      NA
    )
    
  )

#extra cols for visualizing
env_vars <- c("log_env_PM25_daily_mean","log_env_PM10_daily_mean","log_env_bc_880_daily_mean","log_env_precip_daily_mean", "env_precip_daily_mean_p99","env_PM25_daily_mean","env_PM10_daily_mean","env_bc_880_daily_mean","env_precip_daily_mean","env_abs_humidity_daily_mean","env_rel_humidity_daily_mean")
```

look at histograms to investigate bins
```{r}
df %>% select(all_of(env_vars)) %>%
  setNames(env_vars) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  mutate(
    variable = factor(variable, levels = env_vars)  
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(
    ~ variable,
    scales = "free",
    ncol = 4,
    labeller = as_labeller(nice_env_labels) 
  ) +
  theme_classic(base_size = 8) +
  theme(
    strip.text = element_text(size = 8),
    axis.text  = element_text(size = 6)
  ) +
  labs(
    title = "Distribution of Environmental Variables",
    x = NULL,
    y = "Count"
  )

```

get correlation matrix
```{r}
# subset and order variables
df_corr <- df %>%
  dplyr::select(all_of(env_vars))

# compute correlation matrix
cor_mat <- cor(df_corr, use = "pairwise.complete.obs")

# apply nice labels
colnames(cor_mat) <- nice_env_labels(colnames(cor_mat))
rownames(cor_mat) <- nice_env_labels(rownames(cor_mat))

corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.cex = 0.7,
  addCoef.col = "black",
  number.cex = 0.5
)
```

plot of temperature and precipitaiton by month
```{r}
envraw$date <- as.Date(envraw$date_neu, format = "%d-%B-%Y")

env_sub <- subset(
  envraw,
  date >= as.Date("2024-07-01") &
  date <= as.Date("2025-08-31")
)

env_sub$month <- format(env_sub$date, "%Y-%m")

monthly_temp <- aggregate(
  Temp_A_LfU_UFP ~ month,
  data = env_sub,
  FUN = mean,
  na.rm = TRUE
)

monthly_precip <- aggregate(
  NI_A_LfU_UFP ~ month,
  data = env_sub,
  FUN = sum,
  na.rm = TRUE
)

monthly <- merge(monthly_temp, monthly_precip, by = "month")
monthly <- monthly[order(monthly$month), ]

monthly$month_date <- as.Date(paste0(monthly$month, "-01"))
monthly$month_label <- format(monthly$month_date, "%b %Y")
monthly$month_id <- seq_len(nrow(monthly))

par(mar = c(8, 4, 4, 5) + 0.1)
scale_factor <- max(monthly$Temp_A_LfU_UFP, na.rm = TRUE) /
                max(monthly$NI_A_LfU_UFP, na.rm = TRUE)

temp_max <- max(monthly$Temp_A_LfU_UFP, na.rm = TRUE)

bp <- barplot(monthly$NI_A_LfU_UFP * scale_factor,
              col = "steelblue",
              border = "steelblue4",
              xaxt = "n",
              #ylim = c(min(monthly$Temp_A_LfU_UFP, na.rm = TRUE),temp_max + 2),
              ylab = "Mean Temperature (°C)",
              #xlab = "Month",
              main = "Monthly Mean Temperature & Total Precipitation")

lines(bp,
      monthly$Temp_A_LfU_UFP,
      col = "#FF506E",
      lwd = 3)

text(x = bp,
     y = par("usr")[3] - 0.08 * diff(par("usr")[3:4]),
     labels = monthly$month_label,
     srt = 45,
     adj = 1,
     xpd = TRUE,
     cex = 0.9)

precip_ticks <- pretty(range(monthly$NI_A_LfU_UFP, na.rm = TRUE))

axis(4,
     at = precip_ticks * scale_factor,
     labels = precip_ticks,
     cex.axis = 0.75)

mtext("Total Precipitation (mm)",
      side = 4,
      line = 3)

# legend
#legend("topright",legend = c("Mean Temperature", "Total Precipitation"),col = c("darkorange2", "steelblue"),lwd = c(3, 10),bty = "n")
```



#2. scatterplot of the three lung funciton outcomes
```{r}
temp_var <- "env_temp"  # hourly temp

p_list <- lapply(names(outcomes), function(o) {
  
  y_var <- outcomes[[o]]$var
  y_title <- outcomes[[o]]$title
  
  ggplot(df, aes_string(x = temp_var, y = y_var)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = TRUE, color = "blue") +
    labs(
      x = "Temperature",
      y = y_title,
      title = paste("Temperature vs\n", y_title)
    ) +
    theme_minimal()
})

# Combine into one figure
wrap_plots(p_list, ncol = 3)
```

```{r}
plot_list <- lapply(1:10, function(pid) {
  
  df_sub <- df %>%
    filter(ID == pid)
  
  p_list <- lapply(names(outcomes), function(o) {
    
    y_var <- outcomes[[o]]$var
    y_title <- outcomes[[o]]$title
    
    ggplot(df_sub, aes_string(x = temp_var, y = y_var)) +
      geom_point(alpha = 0.4) +
      geom_smooth(method = "loess", se = FALSE, color = "blue") +
      labs(
        x = "Temperature",
        y = y_title,
        title = paste("ID", pid, "-", y_title)
      ) +
      theme_minimal()
  })
  
  wrap_plots(p_list, ncol = 3)
})

# Display first one

for (i in seq_along(plot_list)) {
  print(plot_list[[i]])
}
```






#3. Exposures & Outcomes models 


#3.1 Daily DLNM

database set-up
```{r}
env_var <- "env_temp_ufp"
#env_var <- "env_precip_ufp"

env_daily <- imp_env %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarise(
    value = mean(.data[[env_var]], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date) %>%
  mutate(row_id = row_number())

```

hyperparameter set up
```{r}
max_lag <- 7

cb_daily <- crossbasis(
  x = env_daily$value,
  lag = max_lag,
  argvar = list(
    fun = "ns",
    knots = quantile(env_daily$value, c(0.05, 0.5, 0.95), na.rm = TRUE)
  ),
  arglag = list(
    fun = "ns",
    #knots = c(1, 3, 5),
    #knots = logknots(max_lag, 3),
    intercept = TRUE
  )
)

env_key <- env_daily %>%
  select(date, row_id, value)
```

```{r}
df_model <- df %>%
  mutate(
    visit_date = as.Date(v2ht_starttime),
    hourofday = lubridate::hour(v2ht_starttime),
    weekofyear = lubridate::week(v2ht_starttime)
  ) %>%
  left_join(env_key, by = c("visit_date" = "date")) %>%
  filter(!is.na(row_id), row_id > max_lag)
```

building the cross visit
```{r}
cb_for_visits <- cb_daily[df_model$row_id, , drop = FALSE]

df_model2 <- bind_cols(
  df_model,
  as.data.frame(cb_for_visits)
)

predictor_cols <- colnames(cb_for_visits)
```

running the model
```{r}
models  <- list()
cp_list <- list()

aic_table <- data.frame(
  outcome = character(),
  AIC = numeric(),
  stringsAsFactors = FALSE
)

for (o in names(outcomes)) {

  y_var <- outcomes[[o]]$var
  y_title <- outcomes[[o]]$title

  fml <- as.formula(
    paste(
      y_var, "~",
      paste(predictor_cols, collapse = " + "),
      "+ ns(weekofyear, df = 4) + hourofday + v2it_smoking + v2it_sex + v2i_panel + v2it_age + v2ht_respinfection + v2ht_cooperation + (1 | ID)"
    )
  )

  fit <- lmer(fml, data = df_model2, REML = TRUE)
  models[[o]] <- fit
  
  aic_table <- rbind(
    aic_table,
    data.frame(
      outcome = o,
      title = y_title,
      AIC = AIC(fit)
    )
  )

  fe <- fixef(fit)
  V  <- as.matrix(vcov(fit))

  beta_cb <- fe[predictor_cols]
  V_cb    <- V[predictor_cols, predictor_cols, drop = FALSE]

  cp <- crosspred(
    basis = cb_daily,
    coef  = beta_cb,
    vcov  = V_cb,
    cen   = 20,
    at    = quantile(env_daily$value,
                     probs = seq(0.01, 0.99, 0.01),
                     na.rm = TRUE)
  )
  
  names(cp)
  summary(cp$allfit)

  summary(cp$allRRfit)

  cp_list[[o]] <- cp

  # Plot
  y_title <- outcomes[[o]]$title
  title <- gsub(" \\(%\\)| \\(mL\\)", "", y_title)

  # Contour plot
  plot(cp, "contour",
       xlab = "Temperature (°C)",
       ylab = "Lag (days)",
       main = paste(y_title, "- DLNM"),
       cex.lab = 1.4,   # axis labels
       cex.main = 1.55,  # title
       cex.axis = 1.2)  # tick labels
  
  # Overall plot
  plot(cp, "overall",
       xlab = "Temperature (°C)",
       ylab = paste0("Δ ", y_title),
       main = paste(title, "Cumulative Effect (0–7 days)"),
       cex.lab = 1.5,
       cex.main = 1.65,
       cex.axis = 1.3)
}

print(aic_table)
```



#daily DLNM stratified by panel
```{r}
models  <- list()
cp_list <- list()

panels <- sort(unique(df_model2$v2i_panel))
cols   <- c( "firebrick", "steelblue", "darkgreen")

for (o in names(outcomes)) {

  y_var   <- outcomes[[o]]$var
  y_title <- gsub(" \\(%\\)| \\(mL\\)", "", outcomes[[o]]$title)

  cat("Running outcome:", o, "\n")

  cp_panel_list <- list()
  model_panel_list <- list()

  cb_range <- quantile(
    env_daily$value,
    probs = seq(0.01, 0.99, 0.01),
    na.rm = TRUE
  )

  for (p in panels) {

    cat("  Panel:", p, "\n")

    dat_panel <- subset(df_model2, v2i_panel == p)

    fml <- as.formula(
      paste(
        y_var, "~",
        paste(predictor_cols, collapse = " + "),
        "+ ns(weekofyear, df = 4)",
        "+ hourofday",
        "+ v2it_smoking",
        "+ v2it_sex",
        "+ v2it_age",
        "+ v2ht_respinfection",
        "+ v2ht_cooperation",
        "+ (1 | ID)"
      )
    )

    fit <- lmer(
      fml,
      data = dat_panel,
      REML = TRUE
    )

    model_panel_list[[as.character(p)]] <- fit

    fe <- fixef(fit)
    V  <- as.matrix(vcov(fit))

    beta_cb <- fe[predictor_cols]
    V_cb <- V[predictor_cols, predictor_cols, drop = FALSE]

    cp <- crosspred(
      basis = cb_daily,
      coef  = beta_cb,
      vcov  = V_cb,
      cen   = 20,
      at    = cb_range
    )

    cp_panel_list[[as.character(p)]] <- cp

    cat(
      "    non-missing allfit:",
      sum(is.finite(cp$allfit)),
      "\n"
    )
  }

  models[[o]] <- model_panel_list
  cp_list[[o]] <- cp_panel_list

  # y limits
  ylim_vals <- range(
    unlist(
      lapply(
        cp_panel_list,
        function(x) c(x$alllow, x$allhigh)
      )
    ),
    na.rm = TRUE
  )
  
  y_title <- outcomes[[o]]$title
  title <- gsub(" \\(%\\)| \\(mL\\)", "", y_title)
  
  plot(
    NULL,
    xlim = range(cb_range),
    ylim = ylim_vals,
    xlab = "Temperature (°C)",
    ylab = paste0("Δ ", y_title),
    main = paste(title, "Cumulative Effect by Panel"),
    cex.lab = 1.5,
    cex.main = 1.6,
    cex.axis = 1.3
  )

# reference line at no effect
abline(h = 0, col = "grey60", lwd = 1.5, lty = 2)

for (i in seq_along(panels)) {
  
  cp <- cp_panel_list[[as.character(panels[i])]]
  
  fill_col <- adjustcolor(cols[i], alpha.f = 0.2)
  
  polygon(
    c(cp$predvar, rev(cp$predvar)),
    c(cp$alllow, rev(cp$allhigh)),
    col = fill_col,
    border = NA
  )
  
  lines(
    cp$predvar,
    cp$allfit,
    col = cols[i],
    lwd = 3
  )
}

  for (i in seq_along(panels)) {
    
    cp <- cp_panel_list[[as.character(panels[i])]]
    
    # semi-transparent fill
    fill_col <- adjustcolor(cols[i], alpha.f = 0.1)
    
    polygon(
      c(cp$predvar, rev(cp$predvar)),
      c(cp$alllow, rev(cp$allhigh)),
      col = fill_col,
      border = NA
    )
    
    lines(
      cp$predvar,
      cp$allfit,
      col = cols[i],
      lwd = 3
    )
  }

  legend_labels <- c("COPD", "Hypertension", "T2DM")
  #legend( "topright", legend = legend_labels,col = cols, lwd = 3, bty = "n")
}
```



#hourly DLNM
Clean imputed environmental series, add a row ID to create the cross-basis and match to the spirometry df
```{r}
env_df <- imp_env %>%
  arrange(datetime) %>%
  distinct(datetime, .keep_all = TRUE) %>%
  mutate(row_id = row_number())
```

building cross-basis
```{r}
env_var <- "env_temp_ufp"
env_var <- "env_humidity_ufp"
max_lag <- 72

cb_hourly <- crossbasis(
  x = env_df[[env_var]],
  lag = max_lag,
  argvar = list(fun = "ns", knots = c(quantile(env_df[[env_var]], c(0.05, 0.5, 0.95)))),
  arglag = list(fun = "ns", knots = logknots(max_lag, 3), intercept = TRUE)
)
```

```{r}
env_key <- env_df %>%
  select(datetime, row_id, all_of(env_var))

df_model <- df %>%
  mutate(
    visit_hour = round_date(as.POSIXct(v2ht_starttime, tz = "UTC"), "hour"),
    hour = hour(visit_hour)
  ) %>%
  left_join(env_key, by = c("visit_hour" = "datetime")) %>%
  filter(!is.na(row_id), row_id > max_lag)
```

```{r}
cb_for_visits <- cb_hourly[df_model$row_id, , drop = FALSE]

df_model2 <- bind_cols(
  df_model,
  as.data.frame(cb_for_visits)
)

df_model2 <- df_model2 %>%
  mutate(month = factor(month(visit_hour)))

predictor_cols <- colnames(cb_for_visits)
```

run dlnm
```{r}
models  <- list()
cp_list <- list()


aic_table <- data.frame(
  outcome = character(),
  AIC = numeric(),
  stringsAsFactors = FALSE
)


for (o in names(outcomes)) {
  
  y_var <- outcomes[[o]]$var
  y_title <- outcomes[[o]]$title
  
  # Model formula 
  fml <- as.formula(
    paste(
      y_var, "~",
      paste(predictor_cols, collapse = " + "),
      "+ ns(weekofyear, df = 4) + ns(env_rel_humidity_daily_mean, df = 3) + hourofday + v2it_smoking + v2it_sex + v2it_age + v2i_panel + v2ht_respinfection + v2ht_cooperation + (1 | ID)"
    )
  )
  
  # Fit model 
  fit <- lmer(fml, data = df_model2, REML = FALSE)
  models[[o]] <- fit
  
  
  aic_table <- rbind(
    aic_table,
    data.frame(
      outcome = o,
      title = y_title,
      AIC = AIC(fit)
    )
  )

  
  # Extract coefficients
  fe <- fixef(fit)
  V  <- as.matrix(vcov(fit))
  
  beta_cb <- fe[predictor_cols]
  V_cb    <- V[predictor_cols, predictor_cols, drop = FALSE]
  
  # Crosspred
  cp <- crosspred(
    basis = cb_hourly,
    coef  = beta_cb,
    vcov  = V_cb,
    #cen   = median(env_df[[env_var]], na.rm = TRUE),
    #cen = quantile(env_df[[env_var]], 0.01, na.rm = TRUE),
    cen = 21,
    #cen = 10,
    at    = quantile(env_df[[env_var]],
                     probs = seq(0.01, 0.99, 0.01),
                     na.rm = TRUE)
  )
  
  cp_list[[o]] <- cp
  
  # Plots
  
  # 1. Contour
  plot(cp, "contour",
       xlab = "Temperature",
       ylab = "Lag (hours)",
       main = paste(y_title, "- DLNM"))
  
  # 2. Cumulative
  plot(cp, "overall",
       xlab = "Temperature",
       ylab = paste0("Δ ", y_title),
       main = paste(y_title, "Cumulative effect (0–72h)"))
}

print(aic_table)
```

plot cummulative effects at extreme temperatures
```{r}
# Define percentiles (1st and 99th)
temp_q <- quantile(env_df[[env_var]],
                   probs = c(0.01, 0.99),
                   na.rm = TRUE)

# Generate crosspred (ensure correct lag window)
cp <- crosspred(
  basis = cb_hourly,
  coef  = beta_cb,
  vcov  = V_cb,
  cen   = 21,              # reference (change if needed)
  at    = seq(min(env_df[[env_var]], na.rm = TRUE),
              max(env_df[[env_var]], na.rm = TRUE),
              by = 0.1),
  cumul = TRUE
)

# Find closest predicted values to 1st and 99th percentiles
low_idx  <- which.min(abs(cp$predvar - temp_q[1]))
high_idx <- which.min(abs(cp$predvar - temp_q[2]))

# Extract cumulative effects and CIs
low_effect  <- cp$allfit[low_idx]
low_lowCI   <- cp$alllow[low_idx]
low_highCI  <- cp$allhigh[low_idx]

high_effect <- cp$allfit[high_idx]
high_lowCI  <- cp$alllow[high_idx]
high_highCI <- cp$allhigh[high_idx]

# Print results
cat(sprintf(
  "Extreme low temperature (%.1f°C): %.1f (95%% CI: %.1f, %.1f)\n",
  temp_q[1], low_effect, low_lowCI, low_highCI
))

cat(sprintf(
  "Extreme high temperature (%.1f°C): %.1f (95%% CI: %.1f, %.1f)\n",
  temp_q[2], high_effect, high_lowCI, high_highCI
))

```





#GAM functions

GAM delta y
- we do not need to do change in environmental conditions (x) per person doesn't make as much sense because it's all the same city
```{r}
make_gam_change_plot_list <- function(
    df,
    outcomes,
    env_vars,
    covariates,
    id_var = "ID"
) {
  
  plot_list <- list()
  i <- 1
  
  df[[id_var]] <- as.factor(df[[id_var]])
  
  for (okey in names(outcomes)) {
    
    y_var   <- outcomes[[okey]]$var
    y_title <- outcomes[[okey]]$title
    
    for (x in env_vars) {
      
      # Clean data
      vars_needed <- c(id_var, y_var, x, covariates, "month_num")
      #df_clean <- na.omit(df[, vars_needed])
      df_clean <- df %>%
        dplyr::select(all_of(vars_needed)) %>%
        tidyr::drop_na()
      
      # Demean outcome only
      df_change <- df_clean %>%
        group_by(.data[[id_var]]) %>%
        mutate(
          delta_y = .data[[y_var]] - mean(.data[[y_var]], na.rm = TRUE)
        )
      
      #GAM: Δy ~ absolute x 
      
      f <- as.formula(
        paste(
          "delta_y ~ s(", x, ", k = 4) +",
          "s(month_num, bs = 'cc', k = 12) +",
          "s(", id_var, ", bs = 're') +",    #still using random intercept here-independence assumption
          paste(covariates, collapse = " + ")
        )
      )
      
      fit <- mgcv::gam(f, data = df_change, method = "REML")
      
      # Prediction grid (absolute x)
      x_seq <- seq(
        quantile(df_change[[x]], 0.01, na.rm = TRUE),
        quantile(df_change[[x]], 0.99, na.rm = TRUE),
        length.out = 200
      )
      
      # base newdata with x
      newdata <- setNames(data.frame(x_seq), x)
      newdata$month_num <- 6.5   # midpoint of 1–12 cycle
      newdata[[id_var]] <- df_change[[id_var]][1]
      
      # add covariates (held constant)
      for (cov in covariates) {
        if (is.numeric(df_change[[cov]])) {
          newdata[[cov]] <- mean(df_change[[cov]], na.rm = TRUE)
        } else {
          # for factors → use most common category
          newdata[[cov]] <- names(sort(table(df_change[[cov]]), decreasing = TRUE))[1]
        }
      }
      
      # Predict
      pred <- predict(
        fit,
        newdata = newdata,
        se.fit = TRUE,
        exclude = paste0("s(", id_var, ")")
      )
      
      newdata$fit   <- pred$fit
      newdata$lower <- pred$fit - 1.96 * pred$se.fit
      newdata$upper <- pred$fit + 1.96 * pred$se.fit
      
      # Percentiles
      x5  <- quantile(df_change[[x]], 0.05, na.rm = TRUE)
      x95 <- quantile(df_change[[x]], 0.95, na.rm = TRUE)
      
      # Outcome-specific limits
      ylim_vals <- switch(
        okey,
        FEV1FVC = c(-2, 2),
        FEV1    = c(-50, 50),
        FVC     = c(-70, 60),
        c(-50, 50)
      )
      
      # Plot
      p <- ggplot() +
        geom_ribbon(
          data = newdata,
          aes(x = .data[[x]], ymin = lower, ymax = upper),
          fill = "lightblue",
          alpha = 0.3
        ) +
        geom_line(
          data = newdata,
          aes(x = .data[[x]], y = fit),
          color = "black",
          linewidth = 0.9
        ) +
        geom_vline(xintercept = x5,  linetype = 2, color = "gray40") +
        geom_vline(xintercept = x95, linetype = 2, color = "gray40") +
        coord_cartesian(ylim = ylim_vals) +
        labs(
          title = nice_env_labels(x),
          y = paste0("Δ ", y_title),
          x = nice_env_labels(x)
        ) +
        theme_classic(base_size = 9) +
        theme(
          plot.title   = element_text(size = 9, hjust = 0.5),
          axis.title.y = element_text(size = 8),
          axis.text    = element_text(size = 5)
        )
      
      plot_list[[i]] <- p
      i <- i + 1
    }
  }
  
  return(plot_list)
}
```


run GAM across different environmental factors
```{r}
#extra rows make formatting better
env_vars <- c("env_temp_daily_mean","env_temp_night_mean", "env_temp_daily_sd","env_temp_daily_min","env_temp_daily_max","env_PM25_daily_mean","env_PM10_daily_mean","env_bc_880_daily_mean","env_precip_daily_mean","env_abs_humidity_daily_mean","env_rel_humidity_daily_mean","env_rel_humidity_daily_mean")

#env_vars <- c("log_env_PM25_daily_mean","log_env_PM10_daily_mean","log_env_bc_880_daily_mean","log_env_precip_daily_mean")
```

```{r}
cov_fixed <- c("v2ht_respinfection","v2it_sex", "v2it_age", "v2i_panel", "v2it_smoking")

plot_list <- make_gam_change_plot_list(
  df               = df,
  outcomes         = outcomes,
  env_vars         = env_vars,
  covariates       = c("v2ht_respinfection", "v2it_sex", "v2it_age", "v2i_panel", "v2it_smoking"),
  id_var           = "ID"
)
```

```{r}
paginate_exposure_plots(
  plot_list = plot_list,
  outcomes  = outcomes,
  env_vars  = env_vars,
  n_cols    = 4
)
```


run GAM across different temperature metrics
```{r}
env_vars <- c("env_temp_last3h","env_temp_last7h","env_temp_daily_mean","env_temp_yesterday","env_temp_last2d","env_temp_last4d","env_temp_last7d","env_temp_last14d","temp_diff_yesterday","temp_diff_weekly", "temp_diff_weekly", "temp_diff_weekly")
```

```{r}
plot_list <- make_gam_change_plot_list(
  df               = df,
  outcomes         = outcomes,
  env_vars         = env_vars,
  covariates       = cov_fixed,
  id_var           = "ID"
)

paginate_exposure_plots(
  plot_list = plot_list,
  outcomes  = outcomes,
  env_vars  = env_vars,
  n_cols    = 4
)
```




#GEE

GEE linear/nonlinear, delta y function
```{r}
make_gee_change_plot_list <- function(
    df,
    outcomes,
    env_vars,
    covariates,
    y_limits_manual,
    id_var = "ID",
    nonlinear = FALSE   
) {
  plot_list <- list()
  results_list <- list()
  i <- 1
  r <- 1
  
  df[[id_var]] <- as.factor(df[[id_var]])
  
  for (okey in names(outcomes)) {
    
    y_var   <- outcomes[[okey]]$var
    y_title <- outcomes[[okey]]$title
    
    for (x in env_vars) {
      
      vars_needed <- c(id_var, y_var, x, covariates, "month_num", "hourofday", "weekofyear")
      
      df_clean <- df[, vars_needed] %>%
        na.omit() %>%
        arrange(.data[[id_var]], month_num) %>%
        group_by(.data[[id_var]]) %>%
        mutate(delta_y = .data[[y_var]] - mean(.data[[y_var]], na.rm = TRUE)) %>%
        ungroup()
      
      #chose linear or nonlinear
      if (nonlinear) {
        #knots <- quantile(df_clean[[x]], c(0.25, 0.5, 0.75), na.rm = TRUE)
        #knots <- unique(knots) 
        #knots_str <- paste(knots, collapse = ", ")
        #x_term <- paste0("ns(", x, ", knots = c(", knots_str, "))")
        x_term <- paste0("ns(", x, ", df = 3)")
        
      } else {
        x_term <- x
      }
      
      # --- model ---
      f <- as.formula(
        paste(
          "delta_y ~",
          x_term,
          #"+ ns(month_num, df = 4) +",
          "+ sin(2*pi*weekofyear/52) + cos(2*pi*weekofyear/52) +",    #(testing looping)
          paste(covariates, collapse = " + ")
        )
      )
      
      fit <- geepack::geeglm(
        formula = f,
        id      = df_clean[[id_var]],
        data    = df_clean,
        family  = gaussian,
        corstr  = "ar1"
      )
      
      if (!nonlinear) {
        summ <- summary(fit)
        coef_row <- summ$coefficients[x, ]
        
        results_list[[r]] <- data.frame(
          outcome  = okey,
          exposure = x,
          beta     = as.numeric(coef_row["Estimate"]),
          se       = as.numeric(coef_row["Std.err"]),
          p_value  = as.numeric(coef_row["Pr(>|W|)"]),
          stringsAsFactors = FALSE
        )
        
        r <- r + 1
      }
      
      # --- Prediction ---
      x_seq <- seq(
        quantile(df_clean[[x]], 0.01, na.rm = TRUE),
        quantile(df_clean[[x]], 0.99, na.rm = TRUE),
        length.out = 100
      )
      
      newdata <- setNames(data.frame(x_seq), x)
      
      for (cv in covariates) {
        newdata[[cv]] <- if (is.numeric(df_clean[[cv]])) {
          median(df_clean[[cv]], na.rm = TRUE)
        } else {
          factor(levels(df_clean[[cv]])[1], levels = levels(df_clean[[cv]]))
        }
      }
      
      newdata$month_num <- median(df_clean$month_num, na.rm = TRUE)
      newdata$weekofyear <- median(df_clean$weekofyear, na.rm = TRUE)
      newdata$hourofday <- factor(
        round(median(df_clean$hourofday, na.rm = TRUE)),
        levels = sort(unique(df_clean$hourofday))
      )
      
      # --- design matrix ---
      X <- model.matrix(delete.response(terms(fit)), newdata)
      beta <- coef(fit)
      vcov_mat <- vcov(fit)
      
      newdata$pred <- as.vector(X %*% beta)
      newdata$se   <- sqrt(diag(X %*% vcov_mat %*% t(X)))
      
      newdata$lower <- newdata$pred - 1.96 * newdata$se
      newdata$upper <- newdata$pred + 1.96 * newdata$se
      
      # center
      ref <- newdata$pred[1]
      newdata$pred  <- newdata$pred  - ref
      newdata$lower <- newdata$lower - ref
      newdata$upper <- newdata$upper - ref
      
      # percentiles
      x5  <- quantile(df_clean[[x]], 0.05, na.rm = TRUE)
      x95 <- quantile(df_clean[[x]], 0.95, na.rm = TRUE)
      
      ylim_vals <- switch(
        okey,
        FEV1FVC = c(-2, 2),
        FEV1    = c(-77, 75),
        FVC     = c(-75, 75),
        c(-50, 50)
      )
      
      p <- ggplot(newdata, aes(x = .data[[x]], y = pred)) +
        geom_ribbon(
          aes(ymin = lower, ymax = upper),
          fill = "lightblue", alpha = 0.4
        ) +
        geom_line(color = "black", linewidth = 0.8) +
        geom_vline(xintercept = x5,  linetype = 2, color = "gray40") +
        geom_vline(xintercept = x95, linetype = 2, color = "gray40") +
        geom_hline(yintercept = 0, linetype = 2, color = "gray40") +
        coord_cartesian(ylim = ylim_vals) +
        labs(
          title = nice_env_labels(x),
          y = paste0("Δ ", y_title),
          x = nice_env_labels(x)
        ) +
        theme_classic(base_size = 7)
      
      plot_list[[i]] <- p
      i <- i + 1
    }
  }
  
  results_df <- if (!nonlinear) {
    do.call(rbind, results_list) %>%
      dplyr::mutate(exposure = nice_env_labels(exposure)) %>%
      dplyr::arrange(exposure, outcome)
  } else {
    NULL
  }
  
  return(list(
    plots = plot_list,
    results = results_df
  ))
}
```


outcome variables
```{r}
env_vars <- c("env_temp_daily_mean","env_temp_daily_sd","env_temp_daily_min","env_temp_daily_max","env_PM25_daily_mean","env_PM10_daily_mean","env_bc_880_daily_mean","env_precip_daily_mean","env_abs_humidity_daily_mean","env_rel_humidity_daily_mean","env_rel_humidity_daily_mean","env_rel_humidity_daily_mean")

env_vars <- c("env_temp_last3h","env_temp_last7h","env_temp_daily_mean","env_temp_yesterday","env_temp_last2d","env_temp_last4d","env_temp_last7d","env_temp_last14d","temp_diff_yesterday","temp_diff_weekly", "temp_diff_weekly", "temp_diff_weekly")
```

```{r}
cov_fixed <- c("v2ht_respinfection")

res <- make_gee_change_plot_list(
  df               = df,
  outcomes         = outcomes,
  env_vars         = env_vars,
  covariates       = cov_fixed,
  y_limits_manual  = y_limits_manual,
  id_var           = "ID",
  nonlinear = TRUE
)
```

#non linear
```{r}
results_df <- res$results %>%
  dplyr::mutate(
    exposure_pretty = nice_env_labels(exposure),
    beta = round(beta, 3),
    se   = round(se, 3),
    p_value = signif(p_value, 3)
  ) %>%
  dplyr::select(
    outcome,
    exposure = exposure_pretty,
    beta,
    se,
    p_value
  ) %>%
  dplyr::arrange(exposure, outcome)
```

```{r}
results_df

plot_list <- res$plots

paginate_exposure_plots(
  plot_list = plot_list,
  outcomes  = outcomes,
  env_vars  = env_vars,
  n_cols    = 4
)
```

#linear
```{r}
cov_fixed <- c("v2ht_respinfection")

res <- make_gee_change_plot_list(
  df               = df,
  outcomes         = outcomes,
  env_vars         = env_vars,
  covariates       = cov_fixed,
  y_limits_manual  = y_limits_manual,
  id_var           = "ID"
)


results_df <- res$results %>%
  dplyr::mutate(
    exposure_pretty = nice_env_labels(exposure),
    beta = round(beta, 3),
    se   = round(se, 3),
    p_value = signif(p_value, 3)
  ) %>%
  dplyr::select(
    outcome,
    exposure = exposure_pretty,
    beta,
    se,
    p_value
  ) %>%
  dplyr::arrange(exposure, outcome)
```

```{r}
results_df

plot_list <- res$plots

paginate_exposure_plots(
  plot_list = plot_list,
  outcomes  = outcomes,
  env_vars  = env_vars,
  n_cols    = 4
)
```




#pooled models (linear regression for each participant ID)
```{r}
iqr_df <- sapply(env_vars, function(x) {
  IQR(df[[x]], na.rm = TRUE)
})

iqr_df <- data.frame(
  exposure = names(iqr_df),
  IQR = as.numeric(iqr_df),
  stringsAsFactors = FALSE
)
```

variables
```{r}
env_vars <- c("env_temp_daily_mean","env_temp_daily_sd","env_temp_daily_min","env_temp_daily_max","env_PM25_daily_mean","env_PM10_daily_mean","env_bc_880_daily_mean","env_precip_daily_mean","env_abs_humidity_daily_mean","env_rel_humidity_daily_mean")

env_vars <- c("env_temp_last3h","env_temp_last7h","env_temp_daily_mean","env_temp_yesterday","env_temp_last2d","env_temp_last4d","env_temp_last7d","env_temp_last14d","temp_diff_yesterday","temp_diff_weekly")
```

```{r}
run_individual_models <- function(
    df,
    outcomes,
    env_vars,
    covariates,
    id_var = "ID",
    nonlinear = FALSE
) {
  
  results_list <- list()
  r <- 1
  
  for (okey in names(outcomes)) {
    
    y_var <- outcomes[[okey]]$var
    
    for (x in env_vars) {
      
      # prepare data
      df_clean <- df %>%
        select(all_of(c(id_var, y_var, x, covariates, "month_num"))) %>%
        na.omit() %>%
        group_by(.data[[id_var]]) %>%
        mutate(delta_y = .data[[y_var]] - mean(.data[[y_var]], na.rm = TRUE)) %>%
        ungroup()
      
      # loop over individuals
      ids <- unique(df_clean[[id_var]])
      
      for (id_val in ids) {
        
        df_id <- df_clean %>%
          filter(.data[[id_var]] == id_val)
        
        # skip if too few observations
        if (nrow(df_id) < 10) next
        
        # exposure term
        if (nonlinear) {
          x_term <- paste0("ns(", x, ", df = 3)")
        } else {
          x_term <- x
        }
        
        
        cov_keep <- covariates[sapply(covariates, function(cv) {
          length(unique(df_id[[cv]])) > 1
        })]

        
        # formula
        
        f <- as.formula(
          paste(
            "delta_y ~",
            x_term,
            "+ sin(2*pi*month_num/12) + cos(2*pi*month_num/12)",
            if (length(cov_keep) > 0) paste("+", paste(cov_keep, collapse = " + ")) else ""
          )
        )

        
        # fit model
        fit <- lm(f, data = df_id)
        
        if (!nonlinear) {
          coef_row <- summary(fit)$coefficients[x, ]
          
          results_list[[r]] <- data.frame(
            ID       = id_val,
            outcome  = okey,
            exposure = x,
            beta     = coef_row["Estimate"],
            se       = coef_row["Std. Error"],
            p_value  = coef_row["Pr(>|t|)"],
            n_obs    = nrow(df_id),
            stringsAsFactors = FALSE
          )
          
          r <- r + 1
        }
      }
    }
  }
  
  results_df <- do.call(rbind, results_list)
  
  # aggregate (mean effect across individuals)
  summary_df <- results_df %>%
    group_by(outcome, exposure) %>%
    summarise(
      mean_beta = mean(beta, na.rm = TRUE),
      sd_beta   = sd(beta, na.rm = TRUE),
      n_ids     = n(),
      .groups = "drop"
    )
  
  return(list(
    individual_results = results_df,
    summary_results = summary_df
  ))
}
```

```{r}
res_indiv <- run_individual_models(
  df          = df,
  outcomes    = outcomes,
  env_vars    = env_vars,
  covariates  = covariates <- c("hourofday", "v2ht_respinfection", "v2ht_cooperation"),
  nonlinear   = FALSE   # or TRUE
)

res_indiv$individual_results
```

```{r}
iv_results <- res_indiv$individual_results %>%
  filter(!is.na(beta), !is.na(se), se > 0) %>%
  mutate(weight = 1 / (se^2)) %>%
  group_by(outcome, exposure) %>%
  summarise(
    pooled_beta = sum(weight * beta) / sum(weight),
    pooled_se   = sqrt(1 / sum(weight)),
    lower_95    = pooled_beta - 1.96 * pooled_se,
    upper_95    = pooled_beta + 1.96 * pooled_se,
    n_ids       = n(),
    .groups = "drop"
  )
```

```{r}
iv_results_iqr <- iv_results %>%
  left_join(iqr_df, by = "exposure") %>%
  mutate(
    beta_IQR  = pooled_beta * IQR,
    lower_IQR = lower_95 * IQR,
    upper_IQR = upper_95 * IQR,
    result_str = sprintf(
      "%.2f (95%% CI: %.2f, %.2f)",
      beta_IQR, lower_IQR, upper_IQR
    )
  ) %>%
  select(outcome, exposure, result_str)
```


```{r}
# Variables of interest
vars_iqr <- c(
  "env_precip_daily_mean",         # Daily average precipitation
  "env_abs_humidity_daily_mean",   # Daily average absolute humidity
  "env_rel_humidity_daily_mean",   # Daily average relative humidity
  "env_temp_last3h",               # Previous 3 hours average temperature
  "env_temp_daily_mean",           # Daily average temperature
  "env_temp_last14d"               # 2-week average temperature
)

# Calculate IQRs
iqr_values <- data.frame(
  Variable = vars_iqr,
  IQR = sapply(vars_iqr, function(v) IQR(df[[v]], na.rm = TRUE))
)

print(iqr_values)
```


histogram of each var
```{r}
ggplot(res_indiv$individual_results, aes(x = beta)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  facet_grid(outcome ~ exposure, scales = "free") +
  theme_classic()
```

```{r}
# create prediction grid for each exposure
plot_df <- iv_results %>%
  group_by(outcome, exposure) %>%
  do({
    x_vals <- seq(-2, 2, length.out = 100)  # standardized scale (or change)

    data.frame(
      x = x_vals,
      effect = .$pooled_beta * x_vals,
      lower  = .$lower_95 * x_vals,
      upper  = .$upper_95 * x_vals,
      outcome = .$outcome,
      exposure = .$exposure
    )
  }) %>%
  ungroup()
```

```{r}
ggplot(plot_df, aes(x = x, y = effect)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "lightblue", alpha = 0.4) +
  geom_line(color = "black", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_grid(outcome ~ exposure, scales = "free") +
  labs(
    title = "Pooled Linear Exposure–Response",
    x = "Exposure (scaled)",
    y = "Change in outcome"
  ) +
  theme_classic()
```
